---
title: "Joint Report with Fresno BHC on Fresno PD Police Stops"
subtitle: "DRAFT"
author: "Catalyst California and Fresno BHC"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    css: "W:\\RDA Team\\R\\cc_brandguide.css"
    toc: yes
    toc_depth: 2
    toc_float: yes
---
  
<link href="https://fonts.googleapis.com/css2?family=Inter:wght@100;200;300;400;500;600;700;800;900&display=swap" rel="stylesheet">
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)


list.of.packages <- c("usethis", "highcharter", "tidyverse", "sf", "leaflet", "sp", "htmltools", "magrittr", "flextable", "lorem",  "htmlwidgets", "vembedr", "knitr")

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

#### Load libraries ####
library(usethis)
library(highcharter)
library(tidyverse)
library(sf)
library(leaflet)
library(sp)
library(htmltools) # for html escape
library(magrittr)
library(flextable)
library(lorem)
library(htmlwidgets)
library(vembedr)
library(knitr)

#### Add database connection ####
source("W:\\RDA Team\\R\\credentials_source.R")

con <- connect_to_db("eci_fresno_ripa")

## Reknit all embeds so git will recognize file changes
# source("report_embeds/run_embeds.R")

peridot <- "#CEEA01"
orange <- "#F25922"

```

*The San Diego Police Department criminalizes communities of color, especially Black people, in Southeast San Diego by routinely labeling them as gang members. This practice inflicts devastating racially biased harms, undermines community safety, and wastes tremendous public dollars.*  


# ACKNOWLEDGEMENTS

Thank you to the people of Southeast San Diego for sharing their stories. Their stories bring to light the inhumane interactions the Southeast community experiences daily with the police. We are incredibly grateful for their courage and honesty. We urge the City to respond to their stories by putting an end to racially biased policing in San Diego.


# CONTEXT


```{r calgang-table, child= 'report_embeds/report_calgangstats_table.Rmd', echo=FALSE}
```


# BRIEF METHODOLOGY

This report evaluates SDPD gang profiling in the City of San Diego by analyzing data collected by SDPD pursuant to the Racial and Identity Profiling Act (“RIPA”) of 2015.  For each stop made by law enforcement, RIPA data includes key information that can be assessed for evidence of racial or identify profiling—including characteristics about the person stopped (e.g., race, gender, and age), the stop location, the length of the stop, the reason for the stop, and the result of the stop. 

Using RIPA data, this report examines gang profiling by using the officer assignment field (under “gang enforcement”) which we find underrepresents total gang enforcement activity being done by SDPD, and stop result filed (under “field interview card”). 


# FINDINGS

In sum, the data for SDPD clearly show racial bias towards people who officers perceive as Black and/or Latinx, and that these groups are disproportionately impacted by gang enforcement activity. Regions with higher concentrations of  Black and Latinx residents, such as Southeast San Diego, have higher stop rates and are subject to greater gang enforcement activity. Furthermore, comparing stop rates to the San Diego population shows that SDPD stops people perceived as Black at rates much higher than their proportion of the population. In addition to racial bias, the data show that SDPD gang profiling activities often undermine safety by inflicting tremendous harm upon Southeast San Diego community members while rarely resulting in the discovery of criminal evidence or an arrest. . Community member stories add more poignancy to the impact of SDPD’s racially biased activities by showing how Black and Latinx people have personally been harmed repeatedly by officers.

## 1. Fresno PD disproportionately stops people and men of color for traffic stops 


### A. People of color disproportionately stopped generally for traffic stop

```{r bubble-chart, child= 'report_embeds/report_stoprates_race_person_bubble.Rmd', echo=FALSE}
```

```{r bubble-chart, child= 'report_embeds/report_stoprates_result_race_person_stacked.Rmd', echo=FALSE}
```

### B.Men of color especially targeted during these stops  


```{r bubble-chart, child= 'report_embeds/report_stoprates_race_gang_person_bubble.Rmd'}
```

### C. People of color targeted across types of traffic stops  


## 2.  These stops do not advance safety and are a means to profile people of color   

### A. Officers use minor reasons to pull over people of color

```{r donut-chart, child= 'report_embeds/report_hitrates_race_person_donut.Rmd'} 
```

```{r stacked-chart, child= 'report_embeds/report_hitrates_race_ficard_person_bubblepop.Rmd'}
```

### B. These stops most commonly result in citations that could be better addressed through non-law enforcement alternatives 

```{r dot-map, child= 'report_embeds/report_hitrates_div_race_ficard_person_density_map.Rmd'}
```

### C. Furthermore, use of citations is biased against people of color 


## 3.  These traffic stops inflict additional harms on people of color, especially men of color. 

### A. They are used to inflict harm of men of color through use of force and searches 

```{r item-chart, child= 'report_embeds/report_timespent_div_ficard_race_person_group.Rmd'}
```


## 4. Fresno PD spends a disproportionate amount of time on traffic stops and wastes further resources based on their racial bias. 

### A. Most of Fresno PD time is spent on traffic stops

### B. They are biased in how they spend their time during these stops

## 5. The City of Fresno annually wastes millions of dollars on Fresno PD’s harmful and biased practice

### A. The City of Fresno spends more money on Fresno PD than on things that promote safety

# RECOMMENDATIONS

1. Alternative investments that actually increase safety
2. Enact the Protect Act
3. Others...


# AUTHORS

**Research and Data Analysis**

Elycia Mulholland Graves, Director, Research & Data Analysis, Catalyst California

Jennifer Zhang, Senior Research & Data Analyst, Catalyst California

David Segovia, Research & Data Analyst, Catalyst California

Hillary Khan, Data Architect, Catalyst California

**Authors**

Elycia Mulholland Graves, Director, Research & Data Analysis, Catalyst California

Chauncee Smith, Associate Director, Reimagine Justice & Safety, Catalyst California

Mitchelle Woodson, Legal Director, Pillars of the Community

Jennifer Zhang, Senior Research & Data Analyst, Catalyst California

**Communications**

Ron Simms, Jr., Associate Director, Communications, Catalyst California

Roxanna Reyes, Senior Digital Communications Associate, Catalyst California

**Video Storytelling**

[List videographer, and other Pillars would like to have acknowledged]


# METHODOLOGY

[link to methodology on GitHub]