---
title: "Joint Report with Fresno BHC on Fresno PD Police Stops"
subtitle: "DRAFT"
author: "Catalyst California and Fresno BHC"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    css: "W:\\RDA Team\\R\\cc_brandguide.css"
    toc: yes
    toc_depth: 2
    toc_float: yes
---
  
<link href="https://fonts.googleapis.com/css2?family=Inter:wght@100;200;300;400;500;600;700;800;900&display=swap" rel="stylesheet">
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)


list.of.packages <- c("usethis", "highcharter", "tidyverse", "sf", "leaflet", "sp", "htmltools", "magrittr", "flextable", "lorem",  "htmlwidgets", "vembedr", "knitr")

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

#### Load libraries ####
library(usethis)
library(highcharter)
library(tidyverse)
library(sf)
library(leaflet)
library(sp)
library(htmltools) # for html escape
library(magrittr)
library(flextable)
library(lorem)
library(htmlwidgets)
library(vembedr)
library(knitr)

#### Add database connection ####
source("W:\\RDA Team\\R\\credentials_source.R")

con <- connect_to_db("eci_fresno_ripa")

## Reknit all embeds so git will recognize file changes
# source("report_embeds/run_embeds.R")

peridot <- "#CEEA01"
orange <- "#F25922"

```

*All people in Fresno deserve to feel safe. However, Fresno PD uses traffic stops to profile people of color. This biased practice does not advance safety rather it wastes time and inflicts harm on BIPOC communities. We need a new vision of community safety and reinvestment to keep our families of color safe and together.*  


# ACKNOWLEDGEMENTS

# CONTEXT

# BRIEF METHODOLOGY

# FINDINGS


## 1. Fresno PD disproportionately stops people and men of color for traffic stops 


### A. People of color disproportionately stopped generally for traffic stop

```{r bubble-chart, child= 'report_embeds/report_traffic_race.Rmd'}
```


### B.Men of color especially targeted during these stops  

```{r bubble-chart, child= 'report_embeds/report_traffic_race_gender.Rmd'}
```


### C. People of color are targeted across all types of traffic stops  

```{r stacked-chart, child= 'report_embeds/report_traffic_type_race.Rmd'}
```

## 2.  These stops do not advance safety and are a means to profile people of color   

### A. Officers use minor reasons to pull over people of color



```{r stacked-chart, child= 'report_embeds/report_traffic_reason_race.Rmd'}
```


### B. These stops most commonly result in citations that could be better addressed through non-law enforcement alternatives 

```{r bubble-chart, child= 'report_embeds/report_traffic_result.Rmd'}
```


```{r table, child= 'report_embeds/report_citation_traffic_type_stop.Rmd'}
```


### C. Furthermore, use of citations is biased against people of color 

```{r bubble-chart, child= 'report_embeds/report_traffic_result_citation_race.Rmd'}
```

```{r bubble-chart, child= 'report_embeds/report_traffic_result_citation_license_race.Rmd'}
```

## 3.  These traffic stops inflict additional harms on people of color, especially men of color. 

### A. Actions taken during traffic stops are biased toward men of color 

```{r stacked-chart, child= 'report_embeds/report_traffic_uof_visual.Rmd'}
```

```{r bar-chart, child= 'report_embeds/report_traffic_detained_visual.Rmd'}
```

## 4. Fresno PD spends a disproportionate amount of time on traffic stops and wastes further resources based on their racial bias. 

### A. Most of Fresno PD time is spent on traffic stops


```{r bar-chart, child= 'report_embeds/report_timespent_reason.Rmd'}
```

### B. They are biased in how they spend their time during these stops


```{r jitter-boxplot, child= 'report_embeds/report_timespent_traffic_race.Rmd'}
```

## 5. The City of Fresno annually wastes millions of dollars on Fresno PD’s harmful and biased practice

### A. The City of Fresno spends more money on Fresno PD than on things that promote safety

# RECOMMENDATIONS

# AUTHORS

**Research and Data Analysis**

Elycia Mulholland Graves, Director, Research & Data Analysis, Catalyst California

Jennifer Zhang, Senior Research & Data Analyst, Catalyst California

Hillary Khan, Data Architect, Catalyst California

Alicia Võ, Research and Data Analyst

# METHODOLOGY

[link to methodology on GitHub]