---
title: "Traffic Stops and Race: An Analysis of Police Stops in Fresno"
author: "Catalyst California in partnership with Fresno Building Healthy Communities"
date: "24 January, 2025"
output:
  html_document:
    toc: yes
    toc_depth: 2
    toc_float: yes
    includes:
      in_header: header.html
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

#### Load libraries ####
library(usethis)
library(highcharter)
library(dplyr)
library(htmltools) # for html escape
library(knitr)

#### Add database connection ####
source("W:\\RDA Team\\R\\credentials_source.R")

con <- connect_to_db("eci_fresno_ripa")

## Reknit all embeds so git will recognize file changes
# source("report_embeds/run_embeds.R")

```

All people in Fresno deserve to feel safe. However, the Fresno Police Department uses racially biased traffic stops to profile people of color, thus denying that security to a significant number of Fresno’s residents. This practice undermines community safety, inflicts harm on Black, Indigenous, and People of Color (BIPOC) communities, and wastes time and public dollars. We need a new vision of community safety and reinvestment to keep our families safe and together.

<br>

> __“For me, safety in my community means: safe spaces for all intersectionalities, less presence of police brutality, access to the lighting in neighborhoods, community drives, more sidewalks, green spaces, access to safe and clean parks, more safety in air and water, justice and safety for immigrants, and access to all resources.” -Fresno community member__    
<br>
__*“Para mí, la seguridad en mi comunidad significa: espacios seguros para todos sin tomar en cuenta su raza, sexo, estado civil, etcétera; menos brutalidad policíaca; más iluminación en las calles de los vecindarios; campañas comunitarias en la comunidad; más aceras, áreas verdes, acceso a parques seguros y limpios; más seguridad en el medio ambiente; justicia y seguridad para los inmigrantes; y acceso a todos los recursos.” -Miembro de la comunidad de Fresno*__

<br>

# ACKNOWLEDGEMENTS

Catalyst California completed this research project in collaboration with Fresno Building Healthy Communities (“FBHC”). It was guided by invaluable input from FBHC’s coalition partners and community members. Their perspectives and lived experiences are the driving force behind the data.

# CONTEXT

The City of Fresno has a complicated history with the Fresno Police Department (“Fresno PD”) and its interaction with the communities it is supposed to protect and serve--especially marginalized communities. The city and its police department have a history of harming communities of color through dangerous traffic policies and practices, use of force against members of the community, and gang profiling.

<br>
Fresno has large Southeast Asian and Latinx populations that feel this harm: Southeast Asian people through gang profiling, and Latinx men through disproportionate detention during traffic stops. The department also continues to stop and harass Fresno youth as they walk to and from school. These youth have limited access to safe spaces across the city.

<br>
Fresno PD says its focus is to increase the quality of life for community members by reducing both violent and nuisance crime, like graffiti and petty larceny, and lowering fear of crime among residents. They say they do this by working with community. But the majority of officer patrol time is spent conducting traffic stops. As a result, communities of color experience constant over-policing, harassment, and physical harm. 

<br>
This report focuses on the harm Fresno PD perpetuates through traffic stops and the real impacts communities face as a result. Traffic stops are a starting point to profiling, detaining, harassing, and harming communities of color, thus denying them the safety they deserve to thrive. 

<br>
The City of Fresno must change its approach to community safety by shifting away from an overreliance on Fresno PD and toward equity and care-centered community safety solutions. Data show Fresno PD disproportionately stops and tickets people of color for minor traffic violations, such as equipment or administrative issues, that do not meaningfully advance community safety. This kind of racial and identity profiling inflicts trauma and economic extraction on BIPOC communities. Fresno PD’s commitment to such unproductive practices wastes more than $200 million public dollars annually.

<br>
The City of Fresno should end racially biased traffic stops, enhance Fresno PD’s accountability and transparency, and reinvest dollars wasted on such profiling into programs that address the root causes of community safety risks. They should also meaningfully listen to Fresno community members, who have a clear picture of what safety means to them.

<br>

> __"Que más [nos] sintamos en confianza con las autoridades de la ley, como, por ejemplo, que tengamos a los agentes de policía como héroes y no que atentan contra nuestras vidas." -Miembro de la comunidad de Fresno__    
<br>
__*"That we can better trust law enforcement; for example, that we can hold police officers as heroes that don’t threaten our lives." -Fresno community member*__
<br>

<br>

> __"Mejor[es] calles. Menos basura, más limpieza. Más tiendas, hospitales, basureros en paradas de autobús." -Miembro de la comunidad de Fresno__    
<br>
__*"Better streets. Less trash and more cleanliness. More stores, hospitals, trash cans at bus stops." -Fresno community member*__

# BRIEF METHODOLOGY

This report evaluates Fresno PD activity by analyzing 2022 data collected pursuant to the Racial and Identity Profiling Act (“RIPA”) of 2015. RIPA requires law enforcement officers to collect and report information on each stop they conduct, including the time and location, why the stop was conducted and what occurred during it, as well as characteristics about the person stopped (e.g., race, gender, and age). This report examines profiling by Fresno PD by analyzing RIPA data on who officers choose to stop, and the actions taken by officers during stops.

<br>
Officers collect RIPA data based on their perceptions. This is the standard approach for collecting and reporting data on racial profiling throughout the United States because racial and identity profiling occurs based on officers’ perceptions, not on how people choose to identify. Thus, data on racial profiling may be inconsistent with other demographic data sources. One of the options that RIPA has for a person’s race is Middle Eastern/South Asian (“MESA”). For this report, we use Southwest Asian or North African, or South Asian (“SWANA/SA”) as a more inclusive and accurate representation of these groups.

<br>
Throughout the report, we pair RIPA data with stories from community members to provide a more holistic perspective on police-community interactions. We collected these community stories during a community meeting in October 2024.

# FINDINGS

Fresno PD does not respond to community needs and thus fails to advance community safety. Residents say that when they attempt to call Fresno PD, they experience long wait times, or dispatchers even fail to connect their calls. Data show that Fresno PD focuses primarily on officer-initiated stops, rather than on responding to calls for service. In 2022, only 5.7% of all people Fresno PD stopped were in response to calls for service, while 94.3% were targeted for officer-initiated stops.^[In 2022, Fresno PD stopped a total of 5,365 people across officer-initiated stops and calls for service--officer-initiated stops included 5,060 people and calls for service 305 people. Multiple people can be stopped in a single stop. There were 5,317 unique stops by Fresno PD in 2022, meaning most Fresno PD stops included a single person rather than multiple individuals.]  

<br>
Not only does Fresno PD primarily conduct officer-initiated stops, but the overwhelming majority of these stops are traffic stops. Specifically, 99.3% of officer-initiated stops are traffic stops that do not contribute to community safety.^[This percentage equals 5,023 people Fresno PD stopped for officer-initiated traffic stops in 2022.] 

<br>
Fresno community members would rather have police respond to what they identify as real safety concerns, such as cases of physical violence on children or domestic violence. Their vision of a safe community is free of over-policing and instead includes safe parks and accessible public transit as well as investments in programs for youth and mental health/housing services. 


## 1. Fresno PD disproportionately stops people of color, especially men, for traffic stops

<br>

> __"[For me safety means being] safe and free to go about one’s business without fear or harassment. Able to walk, drive, bike to work, school, or church without worry." -Fresno community member__    
<br>
__*"[Para mí la seguridad significa] sentirme seguro y libre para manejar mis asuntos sin temor ni acoso. Poder caminar, manejar o ir en bicicleta al trabajo, a la iglesia o a la escuela sin preocupaciones." -Miembro de la comunidad de Fresno*__

<br>
Ninety-nine percent of all officer-initiated stops by Fresno PD involved an alleged traffic violation. Officers disproportionately targeted people of color, especially men, during these stops.

<br>

### A. People of color are disproportionately subjected to traffic stops

Fresno community members report feeling targeted by police based on their race, apparel and even gender, often without any evidence of crime. The data echoes this.

<br>
Approximately half (50.5%) of people living in the city are Latinx, and a quarter (25.4%) are White. Comparatively small percentages (6.2% and 6.1%, respectively) are Black and SWANA/SA. Yet Fresno PD disproportionately stops Black and SWANA/SA people for traffic reasons. 

<br>
For every 1,000 Black people who live in the city, Fresno PD stopped 16.7 people they perceived as Black in this category. Meanwhile, Fresno PD only stopped 9.6 people perceived as White for every 1,000 White residents.


<br>

> __"[I’ve] been targeted by the simplest things. Like by the way we are dressed. By our tattoos. Color of our skin. The company we are around with. As well as what part of town you’re in, you can be simply be a target for police." -Fresno community member__    
<br>
__*"Me han singularizado por las cosas más simples. Como por la forma en que estamos vestidos. Por nuestros tatuajes. Color de nuestra piel. La gente con quien andamos. Además de en qué parte de la ciudad nos encontramos, simplemente puedes ser un blanco para la policía." -Miembro de la comunidad de Fresno*__

<br>
```{r rdaBubblepop-chart, child= 'report_embeds/report_traffic_race.Rmd'}
```

In addition, community members report that Fresno PD uses many racially biased stops to label people of color—especially Latinx and Southeast Asian people—as gang members. Also known as gang profiling, this practice has devastating consequences. For example, a gang member classification can subject a person to longer periods in jail or prison. It can also “negatively impact people facing relatively low charges because the threat of far more lengthy imprisonment affects plea bargaining, charging decisions, juror perspectives, and other factors in criminal cases. For immigrant communities, this often includes potential deportation.”^[Catalyst California and Pillars of the Community. End Gang Profiling in Southeast San Diego: Data and Stories from Community Members. 2024. https://www.catalystcalifornia.org/campaign-tools/maps-and-data/end-gang-profiling-in-southeast-san-diego-data-and-stories-from-community-members.]

<br>

> __"They assumed that several of my cousins were gang members because of their appearance." -Fresno community member__    
<br>
__*"Asumieron que varios de mis primos eran pandilleros por su apariencia." -Miembro de la comunidad de Fresno*__

<br>

> __"Yes, I personally have been targeted by the police in a routine traffic stop. I was told to get out of my vehicle for suspicion of being involved with gang activity, when at the time, I was wearing a Fresno state sweatshirt and was in college." -Fresno community member__    
<br>
__*"Sí, yo personalmente he sido singularizado por la policía durante un paro de tránsito. Me ordenaron que saliera de mi auto por sospecha de estar involucrado con pandillas. En ese momento yo llevaba puesto un suéter de la universidad de Fresno State, y yo era estudiante." -Miembro de la comunidad de Fresno*__

<br>

### B. Men of color are especially targeted during these stops  

Out of all traffic stops Fresno PD conducted for each racial group, officers stopped people they perceived as male at higher rates than people perceived as female for all groups other than White. The disparity in traffic stops between males and females is highest among American Indian/Alaska Native (AIAN) and SWANA/SA people. Out of all SWANA/SA people Fresno PD stopped for traffic reasons, more than three of four (76.3%) were male.

<br>
```{r rdaGroupedbar-chart, child= 'report_embeds/report_traffic_race_gender.Rmd'}
```

<br>

### C. People of color are targeted across all types of traffic stops  

There are three general categories of traffic violations: 1) equipment 2) non-moving (i.e., administrative) and 3) moving. Equipment violations include things like tinted windows or a bumper in need of repair. Non-moving, or administrative, violations include things like registration or other paperwork issues. Moving traffic stops include violations like speeding or making an illegal turn. Most traffic stops made by Fresno PD are for moving violations. But across all three types of traffic stops, Fresno PD disproportionately targets people of color.

<br>
For every 1,000 Black people living in the city, Fresno PD stopped four Black people for a non-moving or equipment traffic stop, specifically 1.9 for non-moving stops and 2.1 for equipment-related stops.

<br>
Similarly, for every 1,000 Latinx residents, Fresno PD stopped 2.5 Latinx people for equipment or non-moving traffic reasons. Conversely, Fresno PD stopped only 1.7 White people for a non-moving or equipment traffic stop out of 1,000 White people who live in Fresno.

<br>
These disparities are greater than disparities observed in moving traffic stops. Fresno PD stopped Black people for non-moving and equipment traffic stops at a rate 2.4 times greater than the rate for White people. Comparatively, they stopped Black people for moving violations at a rate 1.6 times greater than that for White people. Here is a look at traffic stops made for alleged non-moving and equipment violations.
 
<br>
```{r stackedbar-chart, child= 'report_embeds/report_traffic_type_race.Rmd'}
```

<br>

## 2. These stops do not advance safety and are used to profile people of color  

A deeper analysis of Fresno PD’s traffic stops shows they serve a function that does not meaningfully improve community safety: profiling residents of color.

<br>
They also pose physical, emotional, and financial consequences for the people stopped. Getting a ticket or having a car impounded has financial repercussions, especially since most people need their vehicle to commute to their job.

<br>
Community members also expressed a growing mistrust towards police that is passed down to younger generations as youth, as passengers in the car, often witness a family member being stopped and harassed. 
 
<br>

> __"Getting our cars impounded. Even getting hurt. Being harassed in front of family and our children. Impacting our children’s perception of law enforcement." -Fresno community member__    
<br>
__*"Nos embargaron los autos, nos lastimaron, fuimos acosados delante de nuestra familia y nuestros hijos, que impacto la percepción que nuestros niños tienen de los oficiales de la ley." -Miembro de la comunidad de Fresno*__

<br>

### A. Officers disproportionately subject people of color to non-safety related stops

While most stops by Fresno PD are for speeding, the department’s next two most common stop reasons are obstructed (or tinted) windows and vehicle registration — equipment and non-moving violations that are not directly related to safety and often indicate racial profiling.^[California Law Revision Commission. Annual Report 2022. Sacramento, CA: California Law Revision Commission, 2022, 30. https://www.clrc.ca.gov/CRPC/Pub/Reports/CRPC_AR2022.pdf.] More specifically, Fresno PD stops Latinx people for obstructed windows three times more frequently than they stop people perceived as White for the same reason, while people perceived as Black are stopped for proof of vehicle registration at higher rates than all other groups.

<br>
When defining what safety meant to them, community members did not mention vehicle equipment-related concerns. However, racial disparities increase when people are stopped for these reasons, and they often amount to pretextual stops that indicate racial profiling. 

<br>
When it comes to traffic violations that may pose safety risks, such as moving violations, the data indicate that people of color are less likely to involved. Specifically, Fresno PD stopped White people for speeding at higher rates than people perceived as Black, Asian, and Latinx. Out of all White people officers stopped for a traffic violation, 66.2% were stopped for speeding compared to 56.5% of Latinx people stopped.

<br>
The discrepancy between this statistic and data showing that officers are more likely to use equipment and non-moving violations to stop people of color indicates that racial bias influences officer decisions in traffic stops.      

<br>
```{r facet-chart, child= 'report_embeds/report_traffic_reason_race.Rmd'}
```

<br>
These data also align with community experiences. Fresno community members say not only that Fresno PD targets people for traffic stops based on a person’s appearance, but also that these targeted stops are being made without any evidence of crime.

<br>

> __"[Safety in my community means to] drive safely without fear of being profiled. Friends and family making it safely home after [a] traffic stop. Not being scared of high-speed chases." -Fresno community member__    
<br>
__*"[La seguridad en mi comunidad significa] sentirme seguro mientras manejo, sin temor a ser discriminado. Que amigos y familiares regresen sanos y salvos a casa después de un paro de tránsito. No temer persecuciones de alta velocidad." -Miembro de la comunidad de Fresno*__

<br>

### B. These stops most commonly result in citations that could be better addressed through non-law enforcement alternatives

Analyzing the results of Fresno PD’s traffic stops further shows how these stops do not meaningfully improve safety. Approximately 80.4% of all traffic stops result in a citation for infraction compared to 0.5% that resulted in an arrest. These citations are almost always unrelated to immediate community safety risks. 

<br>
```{r rdaBar-chart, child= 'report_embeds/report_traffic_result_stop.Rmd'}
```

<br>

### C. Fresno PD’s ticketing practices are racially biased  

Fresno PD is more likely to ticket people of color for traffic violations that pose little to no safety risk. For example, Latinx people were three times more likely to get a citation for an obstructed window compared to White people. And nearly 75% of citations given for no vehicle registration were given to people of color. As explained in the recommendations, these citations inflict financial costs on communities for concerns that could be better addressed through non-law enforcement alternatives. Comparatively, officers are more likely to ticket White people for speeding. Approximately, 66.3% of all citations officers issued to White people were for speeding, compared to 51.3% of citations issued to Latinx people. In other words, tickets for equipment and non-moving violations that pose little safety risk inflict greater burdens on people of color.

<br>
```{r facet-chart, child= 'report_embeds/report_traffic_result_citation_race.Rmd'}
```

<br>
The chart below focuses specifically on the people Fresno PD stopped for traffic reasons and cited for driving without a license. This citation can indicate that Fresno PD is profiling immigrants. Community members expressed that Fresno PD targets Latinx people and harasses them to produce their license to determine whether they are undocumented. There are also reports that local law enforcement works closely with Immigration and Customs Enforcement (ICE) to target people for deportation. Fresno PD stopped 273 people perceived as Latinx, Black, Asian, or SWANA/SA and ticketed them for driving without a license, compared to only 40 White people. 

<br>
```{r rdaStackedbar-chart, child= 'report_embeds/report_traffic_result_citation_license_race.Rmd'}
```

<br>

> __"Harmed by traffic stops—targeting of people of color, especially youth. Poor looking vehicles. Undocumented and unlicensed drivers because of economic situations and lack of adequate public transportation." -Fresno community member__    
<br>
__*"Perjudicados por los controles de tránsito, dirigidos a las personas de color, especialmente a los jóvenes, vehículos con aspecto humilde, conductores indocumentados y sin licencia por situaciones económicas y falta de transporte público adecuado." -Miembro de la comunidad de Fresno*__

<br>

## 3. These traffic stops inflict additional harms on people of color, especially men

Fresno PD’s use of traffic stops also causes physical harm and trauma. During traffic stops, officers may choose to take additional actions against the people they stop. They may physically remove people from their vehicles, use weapons or other devices against them to cause physical harm, or detain them. These actions result in long-term physical and mental harm/trauma, as well as more community mistrust and fear of law enforcement. 

<br>

> __"The police tend to go straight to physical force when you don’t understand the situation. They’re aggressive." -Fresno community member__    
<br>
__*"La policía tiende a recurrir directamente a la violencia física cuando uno no entiende la situación. Son agresivos." -Miembro de la comunidad de Fresno*__

<br>

### A. Actions taken during traffic stops are biased against men of color 

We analyzed all instances in which officers used force against people during stops made for suspected traffic reasons. Uses of force include any stop where an officer took at least one of the following actions: baton or other impact weapon used, canine bit or held a person, chemical spray uses, electronic device used, firearm pointed at a person, firearm discharged or used, person physically removed from vehicle, physical or vehicle contact, impact projectile discharged or used. 

<br>
Community member experiences with Fresno PD strongly suggest that the reported use of force in the RIPA data is undercounted. Several community members spoke about how they or their loved ones endured officers’ use of force during traffic stops. Quarterly reports from the Office of Independent Review also indicate that during 2022 there were at least seven officer-involved shootings and 10 civilian complaints of unreasonable force.^[City of Fresno. 2022 Second Quarter Report. Fresno, CA: City of Fresno, 2022. https://www.fresno.gov/wp-content/uploads/2023/01/2022-2nd-Quarter-Report.pdf.] ^[City of Fresno. 2022 Fourth Quarter Report. Fresno, CA: City of Fresno, 2022. https://www.fresno.gov/wp-content/uploads/2023/04/2022-4th-Quarter-Report-003.pdf.]

<br>

> __"Fui parado por la policía injustamente al igual que mi amigo. Nos bajaron de mi pickup a punta de pistola y no era un agente, eran siete policías. Nos chequearon en la computadora y no encontraron nada de nosotros y nos dejaron ir. ¡Siete policías apuntando con sus armas, y nos esposaron!" -Miembro de la comunidad de Fresno__    
<br>
__*"My friend and I were unfairly stopped by the police. They made us get out of my pickup [truck] at gunpoint and it wasn’t a single officer, they were seven. They checked us out on the computer and found nothing on us and let us go. Seven police officers pointing their guns, and they handcuffed us!" -Fresno community member*__

<br>
People perceived as males of color comprised 78.6% of all people against whom officers used force according to RIPA data. Latinx and Asian men comprised most of these instances of force. During officer-initiated stops, officers used force exclusively during stops they started for suspected traffic reasons. Officers initiated half of these stops for reasons that did not indicate an imminent threat to community or traffic safety. These included no vehicle or trailer registration, parking violation, bike headlight violation, and license plates displayed incorrectly.

<br>
```{r rdaStackedbar-chart, child= 'report_embeds/report_traffic_uof_visual.Rmd'}
```

<br>

> __"I have been stopped at gunpoint by multiple officers at once. My old roommate was consistently stopped by police, taken behind Romains and beat. He was eventually shot and killed by an officer." -Fresno community member__    
<br>
__*"Varios agentes de policía a la vez me han detenido a punta de pistola. La policía detenía constantemente a mi antiguo compañero de cuarto, lo llevaba detrás de Romain’s y lo golpeaba. Finalmente, un oficial lo mató a tiros." -Miembro de la comunidad de Fresno*__

<br>
One of the most common actions Fresno PD officers decide to take during a stop is detaining a person. An officer may detain a person if they suspect a crime and decide to investigate and question the person. One in two people (49.1%) officers detained during officer-initiated traffic stops were Latinx men. This exceeds their share of the Fresno population---comprising just  25.6% of the population. Similarly, Black men make up only 2.9% of the Fresno population but represented 11.3% of people officers decided to detain. 

<br>
And while Fresno PD is more likely to detain Latinx and Black men during officer-initiated traffic stops compared to White men, officers are no more likely to find cause to arrest them.

<br>
Officers arrested 19.2% of Latinx men and just 8.3% of Black men they detained, versus 20% of White men. Comparatively, officers were more likely to find no cause or evidence of a crime among Latinx men they detained. More than half (51.9%) of Latinx men they detained were released with a warning or no action. These findings show racial bias among Fresno PD practices. 
 
<br>
```{r rdaBar-chart, child= 'report_embeds/report_traffic_detained_visual.Rmd'}
```

<br>

## 4. Fresno PD spends a disproportionate amount of time on traffic stops and wastes further resources based on their racial bias

Analyzing how officers spend their time on stops shows that this time rarely contributes to community safety. Community members’ stories about their encounters with police reflect this finding.

<br>

> __"Para mí, seguridad en mi comunidad significa protección para mi familia, estar seguros sin peligros." -Miembro de la comunidad de Fresno__    
<br>
__*"For me, safety in my community means protection for my family, being safe without danger." -Fresno community member*__

<br>

### A. Nearly all Fresno PD patrol time is spent on traffic stops 

Of all the time they spent on officer-initiated stops, officers spent 97.8% of minutes on stops for traffic violations compared to just 0.4% of their time on stops for outstanding arrests (meaning an officer conducted the stop because of a warrant for the person’s arrest).
 
<br>
Meanwhile, Fresno residents say they do not feel safe in community spaces. They report that police do not address significant threats to safety, such as domestic violence. Their attempts to reach Fresno PD result in long wait times or calls that dispatchers do not put through to police. All of this contributes to the feeling Fresno PD does not make their community safer and instead exists to harass BIPOC people. 

 
<br>
```{r rdaStackedbar-chart, child= 'report_embeds/report_timespent_reason.Rmd'}
```

<br>

### B. Fresno PD subjects people of color to stops for longer periods than White people

Officers show bias in how they spend their time during traffic stops. Controlling for the stop type and result, officers show more variation in how much time they spend on stopping a person of color compared to a White person. The chart below shows the time officers spent on every person they stopped for traffic reasons where they issued a ticket or warning or where they decided to take no action. 

<br>
While 23% of stops of White people took Fresno PD more than eight minutes, 31% of stops of people of color took officers more than eight minutes.^[The average time Fresno PD spent stopping a White person for traffic reasons that ended with a ticket, warning, or no action was 7.7 minutes compared to 8.2 minutes on equivalent stops of people of color. The data is skewed toward the higher quartiles. The 75th percentile for time spent on these stops of White people was 8 minutes compared to 10 minutes for people of color. The range in time spent on these stops of White people is 3 to 88.3 minutes compared to 2.7 to 120 minutes for people of color.] Whether a person was White or of color made a difference in how much time officers spent on the stop. This relationship persists when controlling for stop reason, result, and actions taken. Specifically, officers spent more time stopping a person perceived as Black and SWANA/SA, compared to an equivalent stop of a person perceived as White. While the 95th percentile for time spent on traffic stops of White people that resulted in a citation, warning, or no action took was 14 minutes, the 95th percentile for Black people was 19 minutes. 
 
<br>
```{r jitter-boxplot, child= 'report_embeds/report_timespent_traffic_race.Rmd'}
```

> __"Safety to me means that my kids can leave the house without fear that they will be harassed by the police." -Fresno community member__    
<br>
__*"Para mí, la seguridad significa que mis hijos pueden salir de casa sin temor a ser acosados por la policía." -Miembro de la comunidad de Fresno*__

<br>

## 5. The City of Fresno annually wastes millions of dollars on Fresno PD’s biased practices

Fresno PD’s practices are not only harmful but also extremely costly. Fresno residents make it clear that they want to see investments in community programs and built environment improvements, such as more streetlights and more pedestrian/bike friendly transit options, instead of continuing to overfund Fresno PD. 

<br>

> __"I would like to see the money going to the police force be spent in a way that is actually helping our community instead of hurting." -Fresno community member__    
<br>
__*"Me gustaría que el dinero que se destina a la policía se gaste de una forma que realmente ayude a nuestra comunidad, en vez de perjudicarla." -Miembro de la comunidad de Fresno*__

<br>

### A. The City of Fresno spends more money on Fresno PD than on things that promote safety

A local budget consists of several funds that capture and categorize revenue, based on the intent of its use. This typically includes: a general fund, a restricted fund, and an enterprise fund. Of these funds, the general fund contains a local jurisdiction’s most flexible dollars, as they can be used to fund various programming and services, as opposed to restricted fund dollars that have to be spent a particular way. 

<br>
The City of Fresno’s proposed budget of Fiscal Year 2024-2025 (FY25) shows that Fresno PD’s total proposed expenditures for this budget cycle are \$288.3 million (including general and special revenues in the operating budget, debt service, and capital funds).^[Mayor’s Proposed Fiscal Year 2025 Budget, City of Fresno, p. B-170. https://www.fresno.gov/wp-content/uploads/2024/05/FY-2025-Mayors-Proposed-Budget.pdf.] This is an increase of over \$2.8 million from the previous fiscal year. In comparison, PARCS’ (Parks, After School, Recreation, and Community Services Department) total proposed expenditures for the same budget cycle are $155.2 million.

<br>
The PARCS department is a better long-term safety investment than spending public dollars on Fresno PD because PARCS’ programs and services address the root causes of safety risks. Specifically, he PARCS’ stated purpose is to “improve the quality of life citywide by providing safe, clean, accessible parks and community centers, offering diverse programs and recreational activities, and fostering meaningful partnerships. In coordination with multiple city departments and community organizations, PARCS oversees 20 community centers, over 80 parks, and multiple special-use facilities.”^[Mayor’s Proposed Fiscal Year 2025 Budget, City of Fresno, p. B-128. https://www.fresno.gov/wp-content/uploads/2024/05/FY-2025-Mayors-Proposed-Budget.pdf.]

<br>
The function of the PARCS department reflects the vision and desires of the community for what effective community safety can look like. Thus, when reimagining community safety, making significant and meaningful investments into this department can promote building out community-care centered infrastructure that keeps communities safe. However, we see that Fresno PD’s budget was 60 percent more than the PARCS budget. 

<br>
In the context of flexible dollars received by each department, Fresno PD receives far more general fund dollars in comparison to PARCS. For Fresno PD’s operating budget, general operating fund dollars make up 89%. However, when looking at PARCS’ operating budget, general fund operating dollars only make up 13.8% of the PARCS’ total budget.

<br>
We see a large disparity in general fund allocation between these two departments. This is important to note because they are both part of Fresno’s safety ecosystem. Rather than continuing to waste public dollars on Fresno PD, the city should prioritize alternatives that can bring about the vision of community safety desired by Fresno residents.

<br>
Both departments’ general fund operating budgets—the amount of flexible dollars needed to pay staff and run the department—reflect the city favoring Fresno PD in funding allocations. For example, while PARCS only received $20.9 million of general operating fund dollars for the FY25 proposed budget cycle, Fresno PD received \$255.9 million—more than 12 times as much. 

<br>
Staffing in these two departments also illustrates how the city of Fresno prioritizes police. Fresno PD has a total of 1,330 budgeted staffing positions, both sworn and civilian, for FY25, while PARCS has only 162 positions. For Fresno PD, the general fund pays for around 1,282 sworn and civilian staff, while PARCS’ general fund dollars pay for around 73 staff. The city spends so much more of its general fund dollars on staffing Fresno PD than it does for PARCS. 

<br>
Overall, how Fresno allocates dollars across these two departments shows a disconnect between decision-makers and residents about what community safety investments best address safety concerns. The police department receives more than 12 times the flexible funding and more than eight times the staff than the department that services parks, after school programs, recreational, and community services. 
 
 
<br>
```{r rdaStackedbar-chart, child= 'report_embeds/report_budget_stacked.Rmd'}
```

<br>

# RECOMMENDATIONS

> __"Invest in mental health services, assist affordable housing, increase green space and community youth centers." -Fresno community member__    
<br>
__*"Invertir en servicios de salud mental, ayudar con viviendas asequibles, aumentar las áreas verdes y centros juveniles comunitarios." -Miembro de la comunidad de Fresno*__

<br>

Residents have long understood that Fresno PD does not create environments that make the community feel safe. In fact, Fresno PD demonstrates the opposite: they see it making BIPOC communities feel unsafe through over-policing and harassment. Interactions with Fresno PD do not inspire reassurance but instead result in higher stop rates, longer interactions, detainments, and uses of force against BIPOC residents.

<br>
To address these issues, we present below a comprehensive set of equity-centered recommendations that advance community safety without the violence and harm of traditional police practices. 


## 1. Reinvest dollars wasted on non-moving and equipment traffic stops to equity centered safety programs    

a. Invest in community organizations and programming that keep communities safe. This requires making significant investments in organizations that help with food and housing insecurity; supporting transportation, built environment, and green space initiatives, including better lights on the streets; and creating more safe public spaces for youth. Youth development services and organizations that combat recidivism and provide diversion and reentry programs are two additional areas for investment. And finally, Fresno needs more investments in organizations that provide health services, including mental, physical, and behavioral health. Ultimately, community members want to fund organizations directly connected to their priorities and how they define safety.

b. Fund community organizations that ensure communities have access to stable and healthy jobs and other resources that give people the income and resources that let them avoid unpaid vehicle registration, unfixed vehicle equipment, and other violations that can lead to harmful pretextual stops. This can look like Fresno investing in community organizations that focus on job expansion and vocational training in the community. It can also include creating fix-it programs that help with vehicle repairs through vouchers and/or partnering with local auto repair shops. This will increase economic activity for local shops while increasing the safe operation of vehicles. The program can be modeled after the [San Joaquin Valley Air District’s](https://ww2.valleyair.org/grants/drive-clean-in-the-san-joaquin/repair/) [Tune-In Tune-Up program](https://valleycan.org/virtual-titu/).

c. Leverage community budget advocacy so community members have agency over how budget dollars are spent. Ultimately, community members want to fund programs and services directly connected to their priorities and how they define safety. Policymakers and decisionmakers should be accountable to those priorities, particularly when it comes to flexible general fund dollars.  

> __"Brindar más ayuda a personas sin hogar, interesarse realmente por sus necesidades." -Miembro de la comunidad de Fresno__    
<br>
__*"Provide more help to unhoused people, take a real interest in their needs." -Fresno community member*__

## 2. Enhance Fresno PD Accountability

a. When a police department or officer harms the community, there must be policies in place that hold them accountable for their acts. In cases of police misconduct, officers must experience real consequences that avoid shifts to different jurisdictions, under-reporting, or no action. Potentially adding an independent commission or oversight office that handles police misconduct is another way to promote more accountability for the department. Consequences for misconduct need to include potential termination but can also take the form of demotions, ensuring policy deters misconduct. The consequences must also be measurable and reportable to the community. 
    
> __"New trainings for officers and how to handle situations. Better turn out times. Meetings within the different communities, to get feedback from community members." -Fresno community member__    
<br>
__*"Más capacitación para los policías de cómo manejar diversas situaciones. Mejorar lapsos de respuesta y asistencia. Reuniones en las diferentes comunidades, para obtener las opiniones de los miembros de la comunidad." -Miembro de la comunidad de Fresno*__

## 3. Increase Fresno PD transparency 

a. Make Fresno PD data transparent and available to the public through public-facing dashboards and other tools, to stop shielding bad cops or procedures within the department. Dashboards can provide data that increases department accountability by providing more details about stops and the people making them. Coordinates of where stops occurred and demographic information on officers making stops are examples of data that a dashboard should provide to increase transparency and foster more public trust.

## 4. Establish care-centered community safety infrastructure

a. Re-imagining safety infrastructure should include ending stops for non-moving and equipment violations that are used to criminalize BIPOC communities. Instead, the city should take a preventative trust-building approach by conducting community trainings on how to address non-moving violations, and by offering incentives and amnesty to encourage residents to voluntarily correct equipment violations. Shifting more resources towards calls for service will also help improve responsiveness to community needs through reduced wait times and fewer dropped calls. Investigating and addressing the causes of long wait times and dropped calls will be a critical part of this shift and building more community trust.
 

# CONCLUSION

All Fresno residents should feel safe. This means the city should invest in upstream drivers of safety, such as accessible and safe parks, reliable and funded youth and mental health services, economic security, and housing services. Currently, residents endure over-policing, low or non-responsiveness to calls for service, and biased traffic stops that lead to profiling and harm for BIPOC communities. Data and community stories show that: 

  1.	Fresno PD disproportionately stops people of color, especially men. 
  2.	Fresno PD patrol activities do not advance safety but facilitate racial profiling.
  3.	Fresno PD traffic stops inflict harms on people of color, especially men. 
  4.	Fresno PD spends a significant amount of time on racially biased stops. 
  5.	The City of Fresno annually wastes millions of dollars on Fresno PD’s harmful and biased policing practices. 

  
The City of Fresno does not now facilitate the community safety that residents desire and deserve. It should address these issues of harm and adopt the comprehensive set of recommendations provided above that builds out an equity-based, care-centered community safety infrastructure that places the needs of the most vulnerable populations at the forefront, to correct historical harms. Until a change in policies and funding allocations occurs that reflects a system to remove harm from communities of color in Fresno, we will continue to see communities suffer and not realize the vision of all people in Fresno feeling safe. 

<br>

> __"Community safety means money going to community, not police. It means prevention, increase in the community services, including mental health therapists, and food, clothing, and shelter. No harassment of our unhoused neighbors or criminalization of youth." -Fresno community member__    
<br>
__*"La seguridad comunitaria significa que el dinero llega a la comunidad, no a la policía. Significa prevención, aumento a los servicios comunitarios, incluyendo asesores de salud mental, y comida, ropa y vivienda. No acosen a nuestros vecinos sin hogar ni criminalicen a la juventud." -Miembro de la comunidad de Fresno*__

# AUTHORS

__CATALYST CALIFORNIA__

__Equity in Community Investments__

* Myanna Khalfani-King, Manager, Equity in Community Investments, Catalyst California
* Michael Nailat, Associate Director

__Research and Data Analysis__

* Elycia Mulholland Graves, Director, Research & Data Analysis, Catalyst California
* Jennifer Zhang, Senior Research & Data Analyst, Catalyst California
* Hillary Khan, Data Architect, Catalyst California
* Alicia Võ, Research and Data Analyst

Thank you to Sandra Soria and Isaac Bushnell for their research and data analyst support early in this project.

__Reimagine Justice and Safety__

Chauncee Smith, Associate Director, Catalyst California

__REPORT CITATION__

Catalyst California and Fresno Building Healthy Communities. "In the Glare of the Blue Lights: An Analysis of Police Stops in Fresno." 2025. https://fresnobhcdev.wpenginepowered.com/ripa-report/[PENDING]. 

# METHODOLOGY

Click [here](https://github.com/catalystcalifornia/fresnoripa/blob/main/report_main.Rmd) for access to our detailed methods.