---
title: ""
output:
  html_document:
    css: "W:\\RDA Team\\R\\cc_charts.css"
---
<!-- link to Google Font -->
<link href="https://fonts.googleapis.com/css2?family=Inter:wght@100;200;300;400;500;600;700;800;900&display=swap" rel="stylesheet">

```{r, include=FALSE}

# set up -----------------------------

library(RPostgreSQL)
source("W:\\RDA Team\\R\\credentials_source.R")

 # library(devtools)
 # remotes::install_github("catalystcalifornia/rdaCharts") 
library(rdaCharts)
library(stringr)
library(htmltools)

#connect to postgres

source("W:\\RDA Team\\R\\credentials_source.R")

con <- connect_to_db("eci_fresno_ripa")

# pull analysis table for visualizing

df<-dbGetQuery(con, "SELECT * FROM report_traffic_result_citation_race")%>%
  mutate(statute_literal_25 = str_to_title(statute_literal_25))%>%
  mutate(statute_literal_25=ifelse(statute_literal_25 %in% "Opr Veh:window Obstructed",
                                   "Obstructed Window",
                               ifelse(statute_literal_25 %in% "Drive W/O License", "Driving Without License",
    ifelse(statute_literal_25 %in% "No Reg:veh/Trailer/Etc", "No Vehicle Registration",
                     ifelse(statute_literal_25 %in% "Use Cellph W/Driv W/O Hfd", "Using Cellphone Without Handheld Device",
                            ifelse(statute_literal_25 %in% "Unsafe Speed:prevail Cond",
                                   "Unsafe Speed while Driving Vehicle", statute_literal_25))))))%>%
  
   filter(grepl("Obstructed Window|No Vehicle Registration",statute_literal_25))%>%
  
   mutate(race = case_when(
    race == 'latinx' ~ "Latinx",
    race =='nh_white' ~ 'White',
    race =='nh_black' ~ 'Black',
    race == 'aian_aoic' ~ 'AIAN',
    race == 'nhpi_aoic' ~ 'NHPI',
    race == 'nh_multiracial' ~ 'Multiracial',
     race == 'sswana_aoic' ~ 'SWANA/SA',
    race=='nh_asian' ~ 'Asian',
    TRUE ~ as.character(race)))%>%
  filter(!grepl("nh_sswana|nh_aian|nh_nhpi", race))

# Aggregate the citations 'use of cellphone w drive without hfd' and 'no hnd hld device w/ drive' as they seem to basically be the same thing

# split up separate dfs for each denominator option

df_race<-df%>%
  filter(denom=="citation_race" )%>%
  group_by(race)%>%
  filter(total>=5)

df_result<-df%>%
  filter(denom=="citation_result")%>%
  group_by(race)%>%
  filter(count>=5)

# visualize DENOM 1: RACE--------------------------

# create caption 

sourcenote<-"Catalyst California calculations based on City of Fresno's Police Stop Data (2022), catalystcalifornia.org, 2024. "

methodnote <- "Analysis for all officer-initiated traffic stops made by officers in 2022. Less than 10 people Fresno PD perceived as NHPI or AIAN were stopped in an officer-initiated traffic stop in 2022. For data masking purposes the data for these groups is suppressed from this chart."

racenote <- "Race/ethnicity: AIAN=American Indian or Alaska Native, NHPI=Native Hawaiian or Pacific Islander, SWANA/SA=Southwest Asian (Middle Eastern) or North African, or South Asian."

caption_text <- paste("<br>", racenote, sourcenote, methodnote)

# create tooltip text

tooltip_text <- "Out of 100 citations Fresno PD gave to people perceived as <b>{point.race}</b>, <b>{point.rate:.1f}%</b> were for <b>{point.statute_literal_25}</b>, or a total of <b>{point.count:.1f}</b> people cited."

```


```{r, echo=FALSE, message=FALSE}

# TESTING FACETED HIGHCHARTER GRAPHS: RACE DENOMINATOR


library(purrr)

list_of_plots <- map(unique(df_race$statute_literal_25), function(x) {
  filtered_data <- df_race %>%
    filter(statute_literal_25 == x) %>%
    arrange(desc(rate))  # Sort by descending y-values
  
  if (nrow(filtered_data) == 0) {
    return(NULL)  # Skip this group if there are no rows
  }

   hchart(filtered_data, "bar", hcaes(x = race, y = rate),
           showInLegend = FALSE,
           , color = "#3A207D") %>% 
    hc_plotOptions(column = list(stacking = "normal")) %>% 
    hc_add_theme(hc_theme_smpl()) %>% 
    hc_subtitle(text = paste0("Citations given for ",x," out of citations received by race")) %>% 
    hc_yAxis(title = list(text = "Percent (%)"))%>%
    hc_xAxis(title = list(text = ""))%>%
   hc_tooltip(pointFormat = tooltip_text)
})

  
browsable(
  tagList(
    tags$h3("Officers Show More Bias in Citations Unrelated to Public Safety"), # Main title
    list_of_plots %>% hw_grid(rowheight = 350)       # Faceted charts
  )
)

```

<div align="left" style="line-height: 12px;"><span style="font-size: 10px;"><br> Catalyst California calculations based on City of Fresno's Police Stop Data (2022), <u>[Catalyst California](https://www.catalystcalifornia.org)</u>, 2024. Analysis for all officer-initiated traffic stops made by officers in 2022 that resulted in a citation for infraction. Less than 5 people Fresno PD perceived as NHPI or AIAN were stopped in an officer-initiated traffic stop that resulted in a citation for infraction in 2022. For data masking purposes the data for these groups is suppressed from this chart.<br>Race/ethnicity: AIAN=American Indian or Alaska Native, NHPI=Native Hawaiian or Pacific Islander, SWANA/SA=Southwest Asian (Middle Eastern) or North African, or South Asian. .</span> </div>
