---
title: "Fresno RIPA Data Project"
subtitle: "Descriptive Analysis"
author: "Catalyst California"
output:
  html_document:
    css: "W:\\RDA Team\\R\\cc_brandguide.css"
    toc: yes
    toc_depth: 3
    toc_float: yes
date: "`r format(Sys.time(), '%d %B, %Y')`"
---
<link href="https://fonts.googleapis.com/css2?family=Inter:wght@100;200;300;400;500;600;700;800;900&display=swap" rel="stylesheet">

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

list.of.packages <- c("usethis", "highcharter", "tidyverse", "sf", "leaflet", "sp", "htmltools", "magrittr", "flextable", "htmlwidgets")

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)


#### Load libraries ####
library(highcharter)
library(tidyverse)
library(sf)
library(leaflet)
library(sp)
library(htmltools) # for html escape
library(magrittr)
library(flextable)
library(htmlwidgets)
library(sf)
library(leaflet)
library(sp)
library(htmltools) # for html escape
library(scales)
library(dplyr)
library(ggplot2)

#add function that better fits the tables to the page    
FitFlextableToPage <- function(ft, pgwidth = 6.5){

  ft_out <- ft %>% autofit()

  ft_out <- width(ft_out, width = dim(ft_out)$widths*pgwidth /(flextable_dim(ft_out)$widths))
  return(ft_out)
    }

#### Add database connection ####

source("W:\\RDA Team\\R\\credentials_source.R")
con <- connect_to_db("eci_fresno_ripa")

#### Read in tables you need from postgres ####

# You can use the RDA Data Dictionary to look up column meta data for any of these tables. The dictionary is structured the same as pgadmin: https://www.healthycity.org/rda-dev/pgdatadictionary/ 

stops<-dbGetQuery(con, "SELECT * FROM rel_stops")
results<-dbGetQuery(con, "SELECT * FROM rel_stops_result")
person<-dbGetQuery(con, "SELECT * FROM rel_persons")
person_reason<-dbGetQuery(con, "SELECT * FROM rel_persons_reason")
age<-dbGetQuery(con, "SELECT * FROM rel_age")
race<-dbGetQuery(con, "SELECT * FROM rel_races_recode") 
person_result<-dbGetQuery(con, "SELECT * FROM rel_persons_result")
actions<-dbGetQuery(con, "SELECT * FROM rel_stops_actions")
reason<-dbGetQuery(con, "SELECT * FROM rel_stops_reason")
person_action<-dbGetQuery(con, "SELECT * FROM rel_persons_actions")
search<-dbGetQuery(con, "SELECT * FROM rel_stops_searches")

# read in offense codes
offense_codes<-dbGetQuery(con, "SELECT * FROM cadoj_ripa_offense_codes_2023")

```

Before engaging in a detailed analysis of Fresno PD patrol activity as outlined in the research plan, we created this descriptive analysis to explore overall trends in the Fresno RIPA data. This helps us determine if our expected findings are likely to show up in the data and allows us to identify any trends we did not expect. We include basic tallies on each of the following:

* [Overview of officer-initiated stops and calls for service](#overview)
* [Demographics of people stopped by Fresno PD](#demographics)
* [A deep dive into traffic stops](#traffic)
    + [Over-all traffic stops by race and gender](#traffic-demo)
    + [Traffic stops by stop-type](#traffic-type)
    + [raffic stops by stop reason](#traffic-reason)
    + [Traffic stops by stop reason](#traffic-result)
* [Results of officer-initiated stops](#results)
    + [Stops resulting in use of force](#useofforce)
    + [Stops resulting in a field interview (FI) card](#ficard)
    + [Stops resulting in referral to US Department of Homeland Security ICE](#ice)
    + [Stops that resulted in a search that resulted in no contraband](#contraband)
* [Time spent on different types of stops](#time)
    + [Time spent on stops by stop reason](#time-reason)
    + [Time spent on traffic stops by stop result](#time-result)
    + [Time spent on traffic stops by race](#time-race)

# Stops overview {#overview}

## Officer-initiated stops versus calls for service

In 2022, there were a __total of 5,317 police stops__ made by Fresno PD. Of those stops, __5,028 were officer-initiated stops__ and __289 were calls for service.__

In 2022, __94.6%__ of Fresno Police Department stops were officer initiated stops while __only 5.4%__ of stops were calls for service.

```{r}
# Use the 'rel_stops' table to calculate the rate and count of officer initiated stops &  calls for service. 

###CALCULATE TOTAL OFFICER INITIATED STOPS##

count_ois <- stops %>%  
  filter(call_for_service == 0) %>% #filters for Officer-Initiated Stops
    select(stop_id) %>% 
       mutate('Total Officer Initiated Stops' = n()) %>%  
         slice(1) %>% 
           select(-stop_id) 

###CALCULATE CALLS FOR SERVICE TOTAL###

count_cfs <- stops %>%
  filter(call_for_service == 1) %>% 
    select(stop_id) %>% 
      mutate('Total Calls for Service' =n()) %>% 
       slice(1) %>% 
           select(-stop_id) 


### CALCULATE THE RATE OF OFFICER INITIATED STOPS AND CALLS FOR SERVICE###

rate_ois <- stops %>%
  group_by(Stop_Type = call_for_service == 0) %>%
    summarise (percent = 100 * n()/ nrow(stops)) %>%
      mutate(percent =round(percent, digits =1),
        Stop_Type = ifelse(Stop_Type, "Officer Initiated Stop", "Calls for Service")) %>%
  rename('Type' = 'Stop_Type')




###BAR CHART to VISUALIZE RATE OF CALLS FOR SERVICE VS OFFICER INITIATED STOP###

rate_ois <- data.frame( #create df to create a visual later
  Type = c("Officer-Initiated Stop", "Calls for Service"),
  percent = c(94.6, 5.4))

ggplot(rate_ois, aes(x = Type, y = percent, fill = Type))+
  geom_bar(stat = "identity") + #to create barplot
  geom_text(aes(label = paste0(round(percent,1), "%")), #add percentage labels within the bars
          position = position_stack(vjust=0.5), #center percentage within the bar
          size = 3,
          color = "white")+ 
  labs(title = "Nearly 95% of All Stops Fresno PD Makes Are Not Responding to Community Calls", 
       subtitle = "Percentage of Police Stops by Stop-Type Made by Fresno PD in 2022",
       x = "", #labels to visual
       y = "") +
  scale_fill_manual(values = c("Officer-Initiated Stop"="#362178",
      "Calls for Service" = "#362178" ))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 0, hjust =0.5), #sets x-axis labels to be horizontal
      legend.position ="none", #removes legend
      plot.title = element_text(hjust=0.5),
      plot.subtitle = element_text(hjust=0.5))+
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=0.5, size=10), #aesthetics of the barchart
        strip.background = element_rect(fill = "#362178"),
        strip.text = element_text(color = "white", size=10, face="bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
panel.background = element_blank())


```


## People stopped for officer-initiated stops

In 2022, Fresno PD stopped a total of __5,060 people__ through officer-initiated stops.

```{r}

# Use the 'rel_persons' AND 'rel_stops' table to calculate total number of people stopped for officer-initiated stops

tot_ois <- left_join(stops, person, by ="stop_id")%>%
  select(call_for_service)%>%
  filter(call_for_service ==0)%>%  
    summarise(count=n())%>%
     ungroup()%>%
       mutate(Total=sum(count))%>%
        rename("Total Number of People Stopped for Officer-Initiated Stops" = count)


```

# Demographics {#demographics}

## Perceived race of people stopped

Fresno PD __mostly stops people they perceive Latinx__ during officer-initiated stops. Over half of people stopped during these encounters are Latinx. For reference, about 50.5% of Fresno City's population is Latinx. As a comparison, __11.3%__ of people they stopped Fresno PD __perceived as Black__ while the __Fresno City's population only comprises of 6.2% Black people__.

```{r}

# Calculate the count and rate of officer-initiated stops by perceived race of person stopped. Use the 'rel_races_recode' table & 'rel_stops'

##CALCULATES COUNT AND RATE FOR OIS BY RACE##

df_race <-  left_join(stops, race, by ="stop_id")%>% #joining tables 
  filter(call_for_service ==0)


df_nh <- df_race %>% ##creates df for nh_race
  select(nh_race)%>% 
  mutate(Total=n())%>%
  group_by(nh_race)%>%
  mutate(Count=n(), Percent = round(Count/Total*100, 1))%>% 
  slice(1)%>%
  arrange(-Percent)

df_sswana <- df_race %>% #creates df for sswana
  mutate(Total=n())%>%
  group_by(sswana_flag)%>%
  mutate(Count=n(), Percent = round(Count/Total*100, 1))%>% 
  slice(1)%>%
  arrange(-Percent)%>%
  ungroup()%>%
  select(sswana_label, Total, Count, Percent)%>%
    rename(nh_race=sswana_label)

df_nhpi <- df_race %>% #creates df for nhpi
  mutate(Total=n())%>%
  group_by(nhpi_flag)%>%
  mutate(Count=n(), Percent = round(Count/Total*100, 1))%>% 
  slice(1)%>%
  arrange(-Percent)%>%
  ungroup()%>%
  select(nhpi_label, Total, Count, Percent)%>%
    rename(nh_race=nhpi_label)

df_aian <- df_race %>% #creates df for aian
  mutate(Total=n())%>%
  group_by(aian_flag)%>%
  mutate(Count=n(), Percent = round(Count/Total*100, 1))%>% 
  slice(1)%>%
  arrange(-Percent)%>%
  ungroup()%>%
  select(aian_label, Total, Count, Percent)%>%
  rename(nh_race=aian_label)
  
final_race <- rbind(df_nh%>%filter(!nh_race %in% c("nh_aian","nh_sswana","nh_nhpi")), 
                    df_sswana%>%filter(nh_race=='sswana'),
                    df_aian%>%filter(nh_race=='aian'),
                    df_nhpi%>%filter(nh_race=='nhpi')) ##combines nh_race and sswana, nhpi, aian

qa<-df_race%>%group_by(nh_race,sswana_flag, aian_flag,nhpi_flag)%>%summarise(count=n())

## Recode for nh_race for visual labels

final_race <- final_race %>%
  mutate(Race = case_when(
    nh_race == 'latinx' ~ "Latinx", 
    nh_race =='nh_white' ~ 'White',
    nh_race =='nh_black' ~ 'Black',
    nh_race == 'nh_asian' ~ 'Asian',
    nh_race == 'aian' ~ 'AIAN',
    nh_race == 'nhpi' ~ 'NHPI',
    nh_race == 'nh_multiracial' ~ 'Multiracial',
     nh_race == 'sswana' ~ 'SWANA/SA', 
    TRUE ~ as.character(nh_race)))
    
##ADDING TOTAL NUMBER TO LABEL##

final_race <- final_race%>%
  mutate(label = paste0("Total: ", Count, "\n", round(Percent, 1),"%"))

##VISUALIZATION FOR RATE OF OIS BY RACE ##

ggplot(final_race, aes(x = reorder(Race, -Percent), y = Percent, fill = Race))+
  geom_bar(stat = "identity") + #to create barplot
  geom_text(aes(label = label), #add percentage labels within the bars
          position = position_stack(vjust=0.5), #center percentage within the bar
          size =3,
          color = "white")+ 
  labs(title = "Latinx Individuals Make Up Half of Officer-Initiated Stops by Fresno PD",
       subtitle = "Percentage of People Stopped During Officer-Initiated Stops by Perceived Race",
       x = "",
       y = "") +
  scale_fill_manual(values = c("#362178", "#362178", "#362178", "#362178", "#362178", "#362178", "#362178","#362178"))+
  theme_bw()+
  theme(plot.title = element_text(hjust =0.5),
        plot.subtitle = element_text(hjust =0.5),
 #sets y-axis labels to be horizontal
      legend.position ="none", #removes legend
      axis.text.y = element_text(angle =0, hjust=1), #Makes y-axis text readable
      axis.text.x = element_text(angle =0, hjust=0.5)) +
  theme(axis.text.x = element_text(angle = 00, vjust = 0.5, hjust=0.5, size=10), #aesthetics of the barchart
        strip.background = element_rect(fill = "#362178"),
        strip.text = element_text(color = "white", size=10, face="bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
panel.background = element_blank())
#Makes x-axis text readable

```


## Perceived age of people stopped 

Individuals __aged 25-34__ experienced the highest numbers of officer-initiated stops by Fresno PD in the year 2022.

```{r}

# Calculate the count and rate of officer-initiated stops by age-brackets of person stopped. Use the 'rel_age' table & 'rel_stops'

age_ois<- left_join(age, stops, by ="stop_id")%>% #Joined tables
  filter(call_for_service ==0)%>% #Filtered for OIS
  select(age_group_re)%>% # Selected table of interest
  mutate(Total=n())%>% #To find total number for OIS 
  group_by(age_group_re)%>% # Grouping by recoded age brackets
  mutate(Count=n(), 
         Percent = round(Count/Total*100, 1))%>% #Create percentages and rounding them to tenths place
  slice(1)%>%
  rename('Age' ='age_group_re')%>%
  arrange(-Percent)

#########VISUALIZATION FOR RATE OF OIS BY AGE GROUP########

ggplot(age_ois, aes(x = Age, y=Percent))+
      geom_bar( stat = "identity", fill="#362178") + #create barchart 
  geom_text(aes( label = paste0(round(Percent, digits=1),"%" ), #round to tenths place within barchart label
                 y= Percent),
            vjust = 1.4, 
            size =3,
            color = "white",
             fontface = "bold")+
    labs(title = "People Ages 25-34 Account for Nearly 35% of Officer-Initiated Stops by Fresno PD",
         subtitle= "Percentage of People Stopped During Officer-Initiated Stops by Perceived Age",
      x = "", #labels
  y = "")+
    scale_x_discrete(labels = wrap_format(10))+
  theme_bw()+
    theme(plot.title = element_text(hjust =0.5, size = 11.5, face="bold"),
          plot.subtitle = element_text(hjust=0.5),
  axis.text.x = element_text(angle = 00, vjust = 0.5, hjust=0.5, size=10), #for label and axis settings
        strip.background = element_rect(fill = "#362178"),
        strip.text = element_text(color = "white", size=10),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
panel.background = element_blank())


```


## Perceived gender of people stopped 

In 2022, people identified as men (males) accounted for __nearly 57%__ of officer-initiated stops by Fresno PD.

```{r}

# Calculate people stopped for officer-initiated stop by perceived gender. Use the table 'rel_persons' & 'rel_stops'

  gender_ois<- left_join(person, stops, by ="stop_id")%>% 
  filter(call_for_service ==0)%>%
  select(g_full)%>% 
  mutate(Total=n())%>%
  group_by(g_full)%>%
  mutate(Count=n(), 
         Percent = round(Count/Total*100, 1))%>%
  slice(1)%>%
  arrange(-Percent)

##RECODING FOR GENDER SO THAT X-AXIS DISPLAYS GENDERS INSTEAD OF 1:5##

gender_ois <- gender_ois %>%
  mutate(Gender = recode(g_full,
                         '1'= 'Male',
                         '2' = 'Female',
                         '3' = "Transgender Man/Boy",
                         '4' = 'Transgender Woman/Girl',
                         '5' = 'Gender Nonconforming'))


#########VISUALIZATION FOR OIS BY PERCEIVED GENDER########

ggplot(gender_ois, aes(x= reorder(Gender, -Percent), y= Percent, fill = Gender))+
      geom_bar( stat = "identity", fill="#362178") + #create barchart 
  geom_text(aes(label = paste0(round(Percent, digits=1),"%"), #round to tenths place within barchart label
                 y= Percent),
            vjust = 1.4, 
            size =3,
            color = "white",
             fontface = "bold")+
    labs(title = "Males Account for Nearly 57% of Officer-Initiated Stops",
         subtitle = "Officer-Initiated Stops by Perceived Gender in 2022",
      x = "", #labels
  y = "")+
    scale_x_discrete(labels = wrap_format(10))+
  theme_bw()+
  theme(plot.title = element_text(hjust =0.5),
         plot.subtitle = element_text(hjust=0.5),
  axis.text.x = element_text(angle = 00, vjust = 0.5, hjust=0.5, size=8), #aesthetics of the barchart
        strip.background = element_rect(fill = "#362178"),
        strip.text = element_text(color = "white", size=10, face="bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
panel.background = element_blank())

```


## Perceived race and age group

In all age groups except for 55-64 and 65 and older, Latinx individuals make up nearly 50% or more of the officer-initiated stops by Fresno PD. For reference, generally Latinx individuals make up 50.5% of the Fresno population. By comparison, Black individuals make up 6.2% of the Fresno population and SWANA/SA individuals 6.1%. __In most age groups, Black individuals make up a higher percent of Fresno PD's officer-initiated stops, beyond their general share of the Fresno population.__

In 2022, White individuals within the age groups of 55-64 and 65 and older were the most frequently stopped during officer-initiated stops by Fresno PD. White individuals may make up a higher share of the population in these age groups in Fresno. Generally, White individuals make up 25.4% of the Fresno population.

```{r}

race_age_ois <- left_join(age, stops, by ="stop_id")%>% 
  left_join(race, by =c("stop_id","person_number"))%>%
  left_join(person%>%mutate(person_number=as.character(person_number)),by =c("stop_id","person_number"))

race_age_ois<-race_age_ois%>%
  filter(call_for_service ==0)%>% #Filtered for OIS
  group_by(age_group_re)%>%
  mutate(total_age=n())%>% #To find total number for each age group
  group_by(age_group_re,nh_race,total_age)%>% # Grouping by all 3 variables and keeping the total per age group recoded
  summarise(Count=n(), # count for the age/race combo
         Percent = round(Count/total_age*100, 1))%>% # percentage for the race combo out of the age category
  slice(1)

##Clean to make Flextable##


race_age_ois <- race_age_ois%>% #Rename column
  rename('Age Group' = age_group_re, 'Total' = total_age)



race_age_ois <- race_age_ois %>% #Recode for race
  mutate(Race = case_when(
    nh_race == 'latinx' ~ "Latinx", 
    nh_race =='nh_white' ~ 'White',
    nh_race =='nh_black' ~ 'Black',
    nh_race == 'nh_asian' ~ 'Asian',
    nh_race == 'nh_aian' ~ 'AIAN',
    nh_race == 'nh_nhpi' ~ 'NHPI',
    nh_race == 'nh_multiracial' ~ 'Multiracial',
      nh_race == 'nh_sswana' ~ 'SSWANA'))

####Deleting nh_race columns to make a cleaner flextable####
race_age_ois <- race_age_ois[, !names(race_age_ois) %in% c("nh_race")]

#####Reordering columns#####
race_age_ois <- race_age_ois %>% 
  select(`Age Group`, Race, everything())

# ##Flextable for Age Group, Race##
#   flextable(race_age_ois) %>% 
#   set_caption('Officer-Initiated Stops in 2022 by Race and Age') %>% 
#   theme_vanilla() %>%
#   width(width=2,unit="in")
  
  ##Visualization for distribution##
ggplot(race_age_ois, aes(x = Race, y = Percent))+
  geom_bar(stat = "identity", fill = "#362178") + #to create barplot
  geom_text(aes(label = paste0(round(Percent,1), "%"),y=Percent+5), #add percentage labels within the bars
          position = position_dodge(2), #center percentage within the bar
          # vjust = "outward",
          size = 3)+
  scale_x_discrete(labels = ~ str_wrap(as.character(.x), 10))+
  labs(title = "Latinx People Make Up Most Stops by Age, but Black People Are Overrepresented", 
       subtitle = "Officer-Initiated Stops in 2022 by Race and Age",
       x = "", #labels to visual
       y = "Percent of Age Group") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 0, hjust =0.5), #sets x-axis labels to be horizontal
      legend.position ="none", #removes legend
      plot.title = element_text(hjust=0.5),
      plot.subtitle = element_text(hjust=0.5))+
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=0.5, size=8), #aesthetics of the barchart
        strip.background = element_rect(fill = "#362178"),
        strip.text = element_text(color = "white", size=10, face="bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
panel.background = element_blank())+
     facet_wrap(. ~ `Age Group`,nrow=4,scales='free_x')

```

## Perceived age by race and gender

In 2022, __people perceived as men (males) were the most frequently stopped group across most age and race combinations__. Latinx men ages 18-34, followed by Latinx women ages 25-34, had the highest number of stops by Fresno PD during officer-initiated stops.

In 2022, __Latinx men ages 17 and under comprised more than 40% of people stopped under the age of 17__. Similarly, Latinx men ages 18-24 made up about 37% of people stopped between the ages of 18-24.

```{r}

# Calculate the distribution of perceived gender/age of people stopped within each perceived racial group. 

age_race_gender_ois<- left_join(age, stops, by ="stop_id")%>% 
  left_join(race, by =c("stop_id","person_number"))%>%
  left_join(person%>%mutate(person_number=as.character(person_number)),by =c("stop_id","person_number"))

age_race_gender_ois_table<-age_race_gender_ois%>%
  filter(call_for_service ==0)%>% #Filtered for OIS
  group_by(age_group_re)%>%
  mutate(total_age=n())%>% #To find total number for each age group
  group_by(age_group_re,nh_race,g_full,total_age)%>% # Grouping by all 3 variables and keeping the total per age group recoded
  summarise(Count=n(), # count for the age/race/gender combo
         Percent = round(Count/total_age*100, 1))%>% # percentage for the race/gender combo out of the age category
  slice(1)

 ##Clean to make Flextable##

age_race_gender_ois_table <- age_race_gender_ois_table %>% #Recode for Race#
  mutate(Race = case_when(
    nh_race == 'latinx' ~ "Latinx", 
    nh_race =='nh_white' ~ 'White',
    nh_race =='nh_black' ~ 'Black',
    nh_race == 'nh_asian' ~ 'Asian',
    nh_race == 'nh_aian' ~ 'AIAN',
    nh_race == 'nh_nhpi' ~ 'NHPI',
    nh_race == 'nh_multiracial' ~ 'Multiracial',
      nh_race == 'nh_sswana' ~ 'SSWANA'))


age_race_gender_ois_table <- age_race_gender_ois_table %>% ##Recode for Gender##
  mutate(Gender = recode(g_full,
                         '1'= 'Male',
                         '2' = 'Female',
                         '3' = "Transgender Man/Boy",
                         '4' = 'Transgender Woman/Girl',
                         '5' = 'Gender Nonconforming'))


age_race_gender_ois_table <- age_race_gender_ois_table%>% ###Renaming columns### 
  rename('Age Group' = age_group_re, 'Total' = total_age )


####Deleting nh_race and  g_full columns to make a cleaner flextable####
age_race_gender_ois_table <- age_race_gender_ois_table[, !names(age_race_gender_ois_table) %in% c("nh_race", "g_full")]

# add percent out of race and age
age_race_gender_ois_table<-age_race_gender_ois_table%>%
  group_by(`Age Group`,Race)%>%
  mutate(total_race=sum(Count))%>%
  ungroup%>%
  mutate(percent_race=Count/total_race*100)

#####Reordering columns#####
age_race_gender_ois_table <- age_race_gender_ois_table%>% 
  select(`Age Group`, Race, Gender, everything())%>%
  filter(Gender=='Male' | Gender=='Female')

# ##Flextable for Age Group, Race, and Gender##
#   flextable(age_race_gender_ois_table) %>% 
#   set_caption('Officer-Initiated Stops Over Time by Age Group, Race, and Gender') %>% 
#   theme_vanilla() %>%
#   width(width=2,unit="in")

ggplot(age_race_gender_ois_table, aes(x = Race, y = Percent,fill=Gender))+
  geom_bar(stat = "identity"
           ) + 
   scale_fill_manual(values = c("#CEC4E2","#362178"))+#to create barplot
  # geom_text(aes(label = paste0(round(Percent,1), "%"),y=Percent+5), #add percentage labels within the bars
  #         # position = position_dodge(2), #center percentage within the bar
  #         # vjust = "outward",
  #         size = 3)+
  scale_x_discrete(labels = ~ str_wrap(as.character(.x), 10))+
  labs(title = "Men Account for Most Stops Across Age and Race Groups", 
       subtitle = "Officer-Initiated Stops in 2022 by Age, Race, and Gender",
       x = "", #labels to visual
       y = "Percent of Age Group") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 0, hjust =0.5), #sets x-axis labels to be horizontal
      plot.title = element_text(hjust=0.5),
      plot.subtitle = element_text(hjust=0.5))+
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=0.5, size=8), #aesthetics of the barchart
        strip.background = element_rect(fill = "#362178"),
        strip.text = element_text(color = "white", size=10, face="bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
panel.background = element_blank())+
     facet_wrap(. ~ `Age Group`,nrow=4,scales='free_x')    

```


# Stop Reasons

The following explores the most common reasons for officer-initiated stops in 2022. __Traffic violations__ are the overwhelming majority of officer-initiated stops that Fresno PD made in 2022, exceeding more than what is seen in other police jurisdictions in California. For this reason, we explore [traffic stops](#traffic) more in depth as its own section. 


## Officer-initiated stops by stop reason

The Fresno Police Department makes an overwhelming majority of their officer-initiated stops for traffic violations. In 2022, __Over 99%__ of people stopped by Fresno PD officers were stopped due to a perceived traffic violation.

```{r}

# Total count and rates of officer-initiated stops by stop reason. Use officer-initiated stops is the denominator. For this use tables: rel_stops & rel_stops_reason 

# Feel free to visualize as a bar graph, table, or whatever you think works best!

##Create df of all officer-initiated stops w/ stop reason variables##
ois_reason <- left_join(stops, person_reason, by ="stop_id")%>%
  filter(call_for_service ==0) %>%
  select(stop_id, reason, everything())

##Create df with rates and counts by stop reason##
count_ois_reason <- ois_reason %>%
  group_by(reason) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  mutate(Percent = Count/sum(Count) * 100) %>%
    arrange(desc(Count)) %>%               # Change column names and value                                                 displays for more legible table
  rename('Reported Stop Reason' = reason, 
         '# of People Stopped' = Count, 
         '% of All People Stopped' = Percent) %>%
     mutate(`% of All People Stopped` = sprintf("%.2f%%", `% of All People Stopped`))

##Flextable for counts and rates of stop reasons##
  flextable(count_ois_reason) %>% 
  set_caption('Reasons for Officer-Initiated Stops by Fresno PD (2022)') %>% 
  theme_vanilla() %>%
  width(width=2,unit="in") %>%
    align(j = 3, align = "right", part = "body") %>%
    align(j = 3, align = "right", part = "header")
```

# Traffic Stops {#traffic}

This section dives deeper into officer-initiated traffic stops made by Fresno PD in 2022, given that traffic stops make up the overwhelming majority of all stops that Fresno PD made in that year. 

## Over-all traffic stops by race and gender {#traffic-demo}

People perceived as Black by Fresno PD are stopped in traffic at disproportionately high rates relative to the total Black population in Fresno city. 

```{r}

# JZ code chunk for prep

### Create traffic violation universe at person level

traffic<-person%>%
  mutate(person_number=as.character(person_number))%>%
  left_join(stops)%>%
  left_join(person_reason, by =c("stop_id","person_number"))%>%
  filter(call_for_service==0 & reason=='Traffic violation')%>% # keep only officer-initiated stops and traffic violations
  select(stop_id, person_number, reason, g_full, reason, traffic_violation_type, rfs_traffic_violation_type.x, rfs_traffic_violation_code.x )
 
# remove the .x from colunm names

names(traffic) <- gsub(x = names(traffic), pattern = "\\.x", replacement = "")  

```


```{r}

# join race data to traffic-person level table

df<-traffic%>%
  left_join(race)
  
# Analyze and visualize race only cut ----------------------

df_race<-df%>%
  mutate(total=n())%>%
  group_by(nh_race)%>%
  mutate(count=n(),
         rate=count/total*100)%>%
  slice(1)%>%
 ungroup()%>%
    mutate(Race = case_when(
    nh_race == 'latinx' ~ "Latinx", 
    nh_race =='nh_white' ~ 'White',
    nh_race =='nh_black' ~ 'Black',
    nh_race == 'nh_asian' ~ 'Asian',
    nh_race == 'nh_aian' ~ 'AIAN',
    nh_race == 'nh_nhpi' ~ 'NHPI',
    nh_race == 'nh_multiracial' ~ 'Multiracial',
     nh_race == 'nh_sswana' ~ 'SWANA/SA', 
    TRUE ~ as.character(nh_race)))%>%
   select(Race, total, count,rate)%>%
arrange(-rate)

# graph

ggplot(df_race, aes(x = reorder(Race, rate), y = rate))+
  geom_bar(stat = "identity", fill = "#362178") + #to create barplot
  geom_text(aes(label = paste0(round(rate,1), "%")), #add percentage labels within the bars
          position = position_stack(vjust=0.5), #center percentage within the bar
          size = 3,
          color = "white")+ 
  scale_x_discrete(labels = ~ str_wrap(as.character(.x), 5))+
  labs(title = "Half of the people Fresno PD stops in traffic are people perceived as Latinx", 
       subtitle = "Perceived race of people stopped in traffic by Fresno PD",
       x = "", #labels to visual
       y = "") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 0, hjust =0.5), #sets x-axis labels to be horizontal
      legend.position ="none", #removes legend
      plot.title = element_text(hjust=0.5),
      plot.subtitle = element_text(hjust=0.5))+
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=0.5, size=10), #aesthetics of the barchart
        strip.background = element_rect(fill = "#362178"),
        strip.text = element_text(color = "white", size=10, face="bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
panel.background = element_blank())+ 
  coord_flip()


```

Within traffic stops of each perceived racial group, those who Fresno PD perceived as male are stopped in traffic at higher rates than those Fresno PD perceived as female across all BIPOC groups. __The gender disparity in traffic stops is highest for individuals perceived SWANA/SA,__ with individuals perceived as male and SWANA/SA being stopped a much higher rate than individuals perceived as female and SWANA/SA. 

Note that the graph below filters for perceived racial groups that had 10 or more people stopped for an officer-initiated traffic stop in 2022. 

```{r}

    
#  Analyze and visualize race+gender cut-----------------------------

gender<-df%>%
    group_by(nh_race)%>%
  mutate(total=n())%>%
  group_by(nh_race, g_full)%>%
  mutate(count=n(),
         rate=count/total*100)%>%
  slice(1)%>%
     mutate(Race = case_when(
    nh_race == 'latinx' ~ "Latinx", 
    nh_race =='nh_white' ~ 'White',
    nh_race =='nh_black' ~ 'Black',
    nh_race == 'nh_asian' ~ 'Asian',
    nh_race == 'nh_aian' ~ 'AIAN',
    nh_race == 'nh_nhpi' ~ 'NHPI',
    nh_race == 'nh_multiracial' ~ 'Multiracial',
     nh_race == 'nh_sswana' ~ 'SWANA/SA', 
    TRUE ~ as.character(nh_race)))%>%
    mutate(Gender = recode(g_full,
                         '1'= 'Male',
                         '2' = 'Female',
                         '3' = "Transgender Man/Boy",
                         '4' = 'Transgender Woman/Girl',
                         '5' = 'Gender Nonconforming'))%>%
  select(Race, Gender, total, count,rate)%>%
  arrange(Race, Gender, -rate)%>%
  filter(total>=10)


# graph

ggplot(gender, aes(x = Gender, y = rate))+
  geom_bar(stat = "identity", fill = "#362178") + #to create barplot
  geom_text(aes(label = paste0(round(rate,1), "%")), #add percentage labels within the bars
          position = position_stack(vjust=0.5), #center percentage within the bar
          size = 3,
          color = "white")+
  scale_x_discrete(labels = ~ str_wrap(as.character(.x), 5))+
  labs(title = "Fresno PD Disproportionately Stops SWANA/SA Men in Traffic",
       subtitle = "Perceived race and gender of people stopped in traffic by Fresno PD",
       x = "", #labels to visual
       y = "") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 0, hjust =0.5), #sets x-axis labels to be horizontal
      legend.position ="none", #removes legend
      plot.title = element_text(hjust=0.5),
      plot.subtitle = element_text(hjust=0.5))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.7, hjust=1, size=10), #aesthetics of the barchart
        strip.background = element_rect(fill = "#362178"),
        strip.text = element_text(color = "white", size=7, face="bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
panel.background = element_blank())+
  facet_wrap(~Race, ncol=4, scales="free")


```

## Traffic stops by stop-type {#traffic-type}

Fresno PD made close to __4,000 traffic stops of moving vehicles in 2022__, which represented more than three-quarters of their total stops for traffic violations. That means the remaining 25% of all traffic stops were made for equipment and non-moving traffic stops. Of the other quarter of vehicles that were stopped, 15% were stopped for equipment violations, while 8% were stopped for non-moving violations.

```{r}

# Total count and rates of traffic stops by each traffic stop type: 1 Moving; 2 Equipment; 3 Nonmoving; Blank. 

#Denominator is out of all officer-initiated traffic stops. 

##Filter out non-traffic violation stops##
tv_reason <- ois_reason %>%
  filter(reason == "Traffic violation")

count_tv_reason <- tv_reason %>%
group_by(traffic_violation_type) %>%
summarise(Count = n(), .groups = 'drop') %>%
mutate(Percent = Count/sum(Count) * 100)

##Create df with count and rates for different traffic stop types##
count_tv_reason <- tv_reason %>%
group_by(traffic_violation_type) %>%
summarise(Count = n(), .groups = 'drop') %>%
mutate(Percent = Count/sum(Count) * 100)

##Create bar chart to visualize counts and rates##
ggplot(count_tv_reason, aes(x = traffic_violation_type, y = Percent, fill = traffic_violation_type)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c("#362178", "#362178", "#362178")) +
  geom_text(aes(label = paste0(Count," people (",round(Percent,1),'%)')),
    vjust = 1.2, color = "white", size = 4) +
  labs(
    title = "Fresno PD Traffic Violation Stops by Type (2022)",
    x = "Traffic Violation Type",
    y = "% of All People Stopped for Traffic Violations") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 2,
                               size = 12,
                               margin = margin(t = 2)),
     axis.text.y = element_text(size = 12),
    legend.position = "none",
    panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
panel.background = element_blank()) +
  scale_y_continuous(limits = c(0, 100))

```

## Traffic stops by stop reason {#traffic-reason}

This analysis focuses the stop reasons specifically for __equipment/non-moving traffic stops__. Over 50% of the non-moving or equipment-related traffic stops made by Fresno PD in 2022 were initiated for vehicles with obstructed windows or improper registration.

```{r}

# This can be a table looking at the top 50 citations within non-moving/equipment traffic stop types. Denominator is non-moving/equipment traffic stops

# Use table 'rel_persons_reason' and the column 'rfs_traffic_violation_code.' You can look up the citation codes here: "W:\Project\ECI\Fresno RIPA\Data\sdcs-look-up-table-2023.v2.xlsx"  

##Filter out moving traffic stops from data set##
tv_nonmove_reason <- tv_reason %>%
  filter(traffic_violation_type != "Moving")

##Create data frame with counts/rates for most frequent citations##
tv_citation_count <- tv_nonmove_reason %>% 
  group_by(rfs_traffic_violation_code) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  mutate(Percent = Count/sum(Count) * 100) %>%
  arrange(desc(Count))

##Create data frame for table visualization with top 10 citations##
tv_citation_flex <- tv_citation_count[1:10, ] %>%
  mutate('Traffic Violation' = case_when(
    rfs_traffic_violation_code == "54571" ~ "Obstructed window/mirrors", 
    rfs_traffic_violation_code == "54657" ~ "Unregistered vehicle",
    rfs_traffic_violation_code == "54644" ~ "License plates improperly displayed",
    rfs_traffic_violation_code == "54099" ~ "Unregistered vehicle in public parking facility",
    rfs_traffic_violation_code == "54109" ~ "Improper vehicle lighting",
    rfs_traffic_violation_code == "54168" ~ "Expired registration",
    rfs_traffic_violation_code == "54116" ~ "Inadequate mufflers",
    rfs_traffic_violation_code == "54141" ~ "Improper bicycle lighting",
    rfs_traffic_violation_code == "54110" ~ "License plate not illuminated",
    rfs_traffic_violation_code == "54191" ~ "Driving without headlights",)) %>% #Adding descriptions of violation indicated by citation code 
  select(rfs_traffic_violation_code, 'Traffic Violation', Count, Percent) %>% #Re-ordering columns
  rename(
    `% of Total Equipment/Non-moving Traffic Stops` = Percent,
    `# of Times Recorded` = Count) %>% # Renaming columns to serve as better table headers
  mutate(`% of Total Equipment/Non-moving Traffic Stops` = sprintf("%.2f%%", `% of Total Equipment/Non-moving Traffic Stops`)) %>% #Improving display of percentages for table
  select(2:4) # Hiding citation codes for table
  
##Create Flex table##
  flextable(tv_citation_flex) %>% 
  set_caption('Reasons for Equipment and Non-moving Traffic Stops by Fresno PD (2022)') %>% 
  theme_vanilla() %>%
  width(width=2,unit="in") %>%
    align(j = 3, align = "right", part = "body") %>%
    align(j = 3, align = "right", part = "header")
  

```

### Traffic stop reasons by race and gender

Within traffic stops that were made by Fresno PD, the most common cited reason is driving at an unsafe speed. However,  people perceived as Black and Latinx are stopped in traffoc for a reason of __'No vehicle registration'__ at higher rates compared to other groups. This is noteworthy as vehicle registration, or lacktherof, does not advance public safety. Traffic stops made for these type of reasons may indicate that profiling is occurring.

Note that the graphs below filter out perceived racial groups where less than 10 people wree stopped in traffic by Fresno PD in 2022. 

```{r}

df0<-traffic%>%
  left_join(race)%>%
  group_by(nh_race)%>%
  mutate(total=n())%>%
  group_by(rfs_traffic_violation_code,nh_race)%>%
  mutate(count=n(),
         rate=count/total*100)%>%
  slice(1)%>%
   mutate(Race = case_when(
    nh_race == 'latinx' ~ "Latinx", 
    nh_race =='nh_white' ~ 'White',
    nh_race =='nh_black' ~ 'Black',
    nh_race == 'nh_asian' ~ 'Asian',
    nh_race == 'nh_aian' ~ 'AIAN',
    nh_race == 'nh_nhpi' ~ 'NHPI',
    nh_race == 'nh_multiracial' ~ 'Multiracial',
     nh_race == 'nh_sswana' ~ 'SWANA/SA', 
    TRUE ~ as.character(nh_race)))%>%
    # mutate(Gender = recode(g_full,
    #                      '1'= 'Male',
    #                      '2' = 'Female',
    #                      '3' = "Transgender Man/Boy",
    #                      '4' = 'Transgender Woman/Girl',
    #                      '5' = 'Gender Nonconforming'))%>%
  select(rfs_traffic_violation_code, Race, total, count,rate)%>%
  group_by(Race)%>%
    arrange(Race, -count)%>%
   slice(1:3)%>%
  left_join(offense_codes, by=c("rfs_traffic_violation_code"="offense_code"))%>%
  select(Race, rfs_traffic_violation_code, statute_literal_25, total, count, rate)%>%
  filter(count>=10) ### Filter total number of people to be a least 10
  
## ggplot-----------


  

ggplot(df0, aes(x = statute_literal_25, y = rate))+
  geom_bar(stat = "identity", fill = "#362178") + #to create barplot
  geom_text(aes(label = paste0(round(rate,1), "%")), #add percentage labels within the bars
          position = position_stack(vjust=0.5), #center percentage within the bar
          size = 3,
          color = "white")+ 
  scale_x_discrete(labels = ~ str_wrap(as.character(.x), .01))+
  labs(title = str_wrap("Top Traffic Citation Reasons by Perceived Race of Person Stopped",110), 
       x = "", #labels to visual
       y = "") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 0, hjust =0.5), #sets x-axis labels to be horizontal
      # legend.position ="none", #removes legend
      plot.title = element_text(hjust=0.5),
      plot.subtitle = element_text(hjust=0.5))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=0.5, size=6), #aesthetics of the barchart
        strip.background = element_rect(fill = "#362178"),
        strip.text = element_text(color = "white", size=8, face="bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
panel.background = element_blank())+
  facet_wrap(. ~ Race, ncol=3, scales = 'free') 

```

### Traffic stops made for driving without license citation

The Fresno Police Department only stopped 9 people in 2022 for driving without a license.

```{r}

# Calculate count of traffic stops that result with a citation for driving without a license. You can look up the citation codes here: "W:\Project\ECI\Fresno RIPA\Data\sdcs-look-up-table-2023.v2.xlsx"

##Get count for how many citations there were for driving w/o license##
noDL_count <- tv_reason %>%
  filter(                         #All citation codes for driving w/o license, using statute VC 12500 (a-d) and VC 1291(a-b)
    rfs_traffic_violation_code %in% c(42127, 54107, 54140, 54113, 54396, 54458, 48106)) %>%
  summarise(count = n())

#print(noDL_count) #count is 9
```

### Equipment traffic stop reasons 

The table below shows the ten most common stop reasons for equipment traffic stops.

```{r}
##Filter all non-Equipment traffic stops from data set##
tv_equip_reason <- tv_reason %>%
  filter(traffic_violation_type == "Equipment")

##Create data frame with counts/rates for most frequent stop reasons and join data with offense code data to add statute description##
tv_equip_count <- tv_equip_reason %>% 
  group_by(rfs_traffic_violation_code) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  mutate(Percent = Count/sum(Count) * 100) %>%
  arrange(desc(Count)) %>%
  left_join(offense_codes, by =c("rfs_traffic_violation_code" = "offense_code")) %>%
  select(rfs_traffic_violation_code, statute_literal_25, Count, Percent, everything())

##Create df for table visualization of top 10 stop reasons##
tv_equip_flex <- tv_equip_count[1:10, ] %>%
  rename(
    `% of Total Equipment Traffic Stops` = Percent,
    `# of Times Recorded` = Count,
    `Traffic Violation` = statute_literal_25) %>% # Renaming columns to serve as better table headers
  mutate(`% of Total Equipment Traffic Stops` = sprintf("%.2f%%", `% of Total Equipment Traffic Stops`)) %>% #Improving display of percentages for table
  select(2:4) # Selecting columns to appear in table

##Create Flextable##
flextable(tv_equip_flex) %>% 
  set_caption('Reasons for Equipment Traffic Stops by Fresno PD (2022)') %>% 
  theme_vanilla() %>%
  width(width=2,unit="in") %>%
    align(j = 3, align = "right", part = "body") %>%
    align(j = 3, align = "right", part = "header")

```

### Non-moving traffic stop reasons

There are heavy inconsistencies in Fresno PD's classification of traffic stops as "non-moving." Of the 10 most common offenses reported as reasons for non-moving traffic stops, six of them also appear in the 10 most common reasons for equipment-related stops, while two also appear in the list of most common reasons for moving stops.

```{r}
##Filter TV to only included non-moving stops##
tv_NMonly_reason <- tv_reason %>%
  filter(traffic_violation_type == "Non-moving")

##Create data frame with counts/rates for most frequent stop reasons and join data with offense code data to add statute description##
tv_NMonly_count <- tv_NMonly_reason %>% 
  group_by(rfs_traffic_violation_code) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  mutate(Percent = Count/sum(Count) * 100) %>%
  arrange(desc(Count)) %>%
  left_join(offense_codes, by =c("rfs_traffic_violation_code" = "offense_code")) %>%
  select(rfs_traffic_violation_code, statute_literal_25, Count, Percent, everything())

##Create df for table visualization of top 10 stop reasons##
tv_NMonly_flex <- tv_NMonly_count[1:10, ] %>%
  rename(
    `% of Total Non-moving Traffic Stops` = Percent,
    `# of Times Recorded` = Count,
    `Traffic Violation` = statute_literal_25) %>% # Renaming columns to serve as better table headers
  mutate(`% of Total Non-moving Traffic Stops` = sprintf("%.2f%%", `% of Total Non-moving Traffic Stops`)) %>% #Improving display of percentages for table
  select(2:4) # Selecting columns to appear in table

##Create Flextable##
flextable(tv_NMonly_flex) %>% 
  set_caption('Reasons for Non-moving Traffic Stops by Fresno PD (2022)') %>% 
  theme_vanilla() %>%
  width(width=2,unit="in") %>%
    align(j = 3, align = "right", part = "body") %>%
    align(j = 3, align = "right", part = "header")

```

### Moving traffic stop reasons

Of the 3,876 moving vehicles stopped for suspected traffic violations by Fresno PD in 2022, __over three-quarters were stopped for speeding__.

```{r}
##Filter TV data to only include moving stops##
tv_move_reason <- tv_reason %>%
  filter(traffic_violation_type == "Moving")

##Create data frame with counts/rates for most frequent stop reasons##
tv_move_count <- tv_move_reason %>% 
  group_by(rfs_traffic_violation_code) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  mutate(Percent = Count/sum(Count) * 100) %>%
  arrange(desc(Count)) %>%
  left_join(offense_codes, by =c("rfs_traffic_violation_code" = "offense_code")) %>%
  select(rfs_traffic_violation_code, statute_literal_25, Count, Percent, everything())

##Create df for table visualization of top 10 stop reasons##
tv_move_flex <- tv_move_count[1:10, ] %>%
  rename(
    `% of Total Moving Traffic Stops` = Percent,
    `# of Times Recorded` = Count,
    `Traffic Violation` = statute_literal_25) %>% # Renaming columns to serve as better table headers
  mutate(`% of Total Moving Traffic Stops` = sprintf("%.2f%%", `% of Total Moving Traffic Stops`)) %>% #Improving display of percentages for table
  select(2:4) # Selecting columns to appear in table

##Create Flextable##
flextable(tv_move_flex) %>% 
  set_caption('Reasons for Moving Traffic Stops by Fresno PD (2022)') %>% 
  theme_vanilla() %>%
  width(width=2,unit="in") %>%
    align(j = 3, align = "right", part = "body") %>%
    align(j = 3, align = "right", part = "header")

```
 

## Traffic stops by result of stop {#traffic-result}

Across all three types of officer-initiated traffic stops (Equipment, Moving, Non-moving), a __citation for infraction__ is the most common stop result, while a __Warning verbal or written__ is within the top three most common stop results across all traffic stops. 

```{r}

# join to person-level traffic universe and analyze

df<-traffic%>%
  left_join(person_result)%>%
  group_by(traffic_violation_type)%>%
  mutate(total=n())%>%
  group_by(traffic_violation_type, stop_result_simple)%>%
  mutate(count=n(),
         rate=count/total*100)%>%
  slice(1)%>%
  select(traffic_violation_type, stop_result_simple, total, count,rate)%>%
  arrange(traffic_violation_type, -rate)%>%
  group_by(traffic_violation_type)%>%
   slice(1:3)

## ggplot-----------
  

ggplot(df, aes(x = stop_result_simple, y = rate))+
  geom_bar(stat = "identity", fill = "#362178") + #to create barplot
  geom_text(aes(label = paste0(round(rate,1), "%")), #add percentage labels within the bars
          position = position_stack(vjust=0.5), #center percentage within the bar
          size = 3,
          color = "white")+ 
  scale_x_discrete(labels = ~ str_wrap(as.character(.x), 10))+
  labs(title = "Most Officer-Initiated Traffic Stops Result in a Citation", 
       subtitle = "Top three most common stop results for officer-initiated traffic stops by traffic-stop type",
       x = "", #labels to visual
       y = "") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 0, hjust =0.5), #sets x-axis labels to be horizontal
      legend.position ="none", #removes legend
      plot.title = element_text(hjust=0.5),
      plot.subtitle = element_text(hjust=0.5))+
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=0.5, size=10), #aesthetics of the barchart
        strip.background = element_rect(fill = "#362178"),
        strip.text = element_text(color = "white", size=10, face="bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
panel.background = element_blank())+
  facet_grid(. ~ traffic_violation_type)




```

The graph below shows, across all traffic stop types (moving, equipment, non-moving), what percentage of stops resulted in a citation for infraction, in-field cite and release, or warning verbal or written across each perceived race group.

Note that the graphs filter out perceived racial groups where less than 10 people were stopped in traffic in 2022. 

Despite being stopped at higher rates than over perceived groups, people Fresno PD perceives as __Latinx and Black__ are stopped in traffic resulting only in a __warning verbal or written__ at higher rates than people perceived as White. 

```{r}

df<-traffic%>%
  left_join(race)%>%
  left_join(person_result)%>%
  filter(stop_result_simple=="in field cite and release"|
          stop_result_simple== "warning verbal or written"|
           stop_result_simple=="citation for infraction")%>%
  group_by(nh_race)%>%
  mutate(total=n())%>%
  group_by(nh_race, stop_result_simple)%>%
  mutate(count=n(),
         rate=count/total*100)%>%
  slice(1)%>%
  ungroup()%>%
   mutate(Race = case_when(
    nh_race == 'latinx' ~ "Latinx", 
    nh_race =='nh_white' ~ 'White',
    nh_race =='nh_black' ~ 'Black',
    nh_race == 'nh_asian' ~ 'Asian',
    nh_race == 'nh_aian' ~ 'AIAN',
    nh_race == 'nh_nhpi' ~ 'NHPI',
    nh_race == 'nh_multiracial' ~ 'Multiracial',
     nh_race == 'nh_sswana' ~ 'SWANA/SA', 
    TRUE ~ as.character(nh_race)))%>%
  select(Race, stop_result_simple, total, count,rate)%>%
  arrange(Race,  -rate)%>%
  filter(total>=10)

# graph


ggplot(df, aes(x = stop_result_simple, y = rate))+
  geom_bar(stat = "identity", fill = "#362178") + #to create barplot
  geom_text(aes(label = paste0(round(rate,1), "%")), #add percentage labels within the bars
          position = position_stack(vjust=0.5), #center percentage within the bar
          size = 3,
          color = "white")+ 
  scale_x_discrete(labels = ~ str_wrap(as.character(.x), 10))+
  labs(title = "Rate of traffic stops by stop result and perceived race",
       x = "", #labels to visual
       y = "") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 0, hjust =0.5), #sets x-axis labels to be horizontal
      legend.position ="none", #removes legend
      plot.title = element_text(hjust=0.5),
      plot.subtitle = element_text(hjust=0.5))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0, hjust=0.5, size=7), #aesthetics of the barchart
        strip.background = element_rect(fill = "#362178"),
        strip.text = element_text(color = "white", size=10, face="bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
panel.background = element_blank())+
  facet_grid(. ~ Race)


```

### Most common citations that resulted from traffic stop

```{r}
# Total count/rate of each citation out of all traffic stops

# read in traffic citations table at person level
citation_result<-dbGetQuery(con, "SELECT * FROM rel_persons_citations")


# join to our traffic universe 

df<-traffic%>%
  left_join(citation_result%>%
              mutate(person_number=as.character(person_number)))

# analyze

df<-df%>%
  select(offense_code)%>%
  mutate(total=n())%>%
  group_by(offense_code)%>%
mutate(count=n(),
       rate=count/total*100)%>%
  slice(1)%>%
  arrange(-rate)%>%
  head(7)%>%
  left_join(offense_codes)%>% # join additional info on offense codes
select(1,6,8,2,3,4)


#create and style flextable

names(df)<-tools::toTitleCase(gsub("_"," ",names(df))) # clean up column names for table

df<-df%>%
  rename("Statute"="Statute Literal 25")

  flextable(df) %>%
colformat_double(j = c("Rate"), digits = 1)%>%
  set_caption('Most Common Citations that resulted from an Officer-Initiated Traffic Stop') %>%
  theme_vanilla() %>%
  width(j=3,2.5,unit="in")


```


### Traffic stops resulting in citations by race and gender

People perceived as Latinx are stopped in traffic and given a citation as a result at the highest rates compared to all other groups. However, __people perceived as Black are disproportionately stopped in traffic and given a citation relative to how many Black people live in Fresno city__. Out of all people stopped in traffic and given a citation, 11.4% were people Fresno PD perceived as Black, yet Black people make up 6.2% of Fresno city's total population. 

The graph below gives more detail into traffic citations given as a result of a traffic stop. It shows the top 3 traffic citation results for perceived race. 

```{r}

## First analyze just traffic stops resulting in a citation by race to see generally which peceived race is receiving the most traffic citations

df<-traffic%>%
  left_join(citation_result%>%
              mutate(person_number=as.character(person_number)))%>%
  left_join(race)

# analyze

# race cut

df_nh<-df%>%
group_by(nh_race)%>%
summarise(count=n(),.groups='drop')%>%
mutate(rate=count/sum(count)*100)%>%
  arrange(nh_race, -rate)

# race/gender cut: I don't think there is a glaring gender disparity here
df_gender<-df%>%
group_by(nh_race, g_full)%>%
summarise(count=n(),.groups='drop')%>%
mutate(rate=count/sum(count)*100)%>%
  arrange(nh_race, -rate)

```

```{r}

df<-traffic%>%
  left_join(citation_result%>%
              mutate(person_number=as.character(person_number)))%>%
  left_join(race)%>%
  group_by(nh_race)%>%
  mutate(total=n())%>%
  group_by(offense_code,nh_race)%>%
  mutate(count=n(),
         rate=count/total*100)%>%
  slice(1)%>%
   mutate(Race = case_when(
    nh_race == 'latinx' ~ "Latinx", 
    nh_race =='nh_white' ~ 'White',
    nh_race =='nh_black' ~ 'Black',
    nh_race == 'nh_asian' ~ 'Asian',
    nh_race == 'nh_aian' ~ 'AIAN',
    nh_race == 'nh_nhpi' ~ 'NHPI',
    nh_race == 'nh_multiracial' ~ 'Multiracial',
     nh_race == 'nh_sswana' ~ 'SWANA/SA', 
    TRUE ~ as.character(nh_race)))%>%
  select(offense_code, Race, total, count,rate)%>%
  group_by(Race)%>%
    arrange(Race, -count)%>%
   slice(1:3)%>%
  left_join(offense_codes, by=c("offense_code"="offense_code"))%>%
  select(Race, offense_code, statute_literal_25, total, count, rate)%>%
  filter(total>=10) # filter for groups where at least 10 people were stopped in traffic


## ggplot-----------
  

ggplot(df, aes(x = statute_literal_25, y = rate))+
  geom_bar(stat = "identity", fill = "#362178") + #to create barplot
  geom_text(aes(label = paste0(round(rate,1), "%")), #add percentage labels within the bars
          position = position_stack(vjust=0.5), #center percentage within the bar
          size = 3,
          color = "white")+ 
  scale_x_discrete(labels = ~ str_wrap(as.character(.x), .1))+
  labs(title = str_wrap("Top 3 Traffic Citation Reasons by Perceived Race of Person Stopped",110), 
       x = "", #labels to visual
       y = "") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 0, hjust =0.5), #sets x-axis labels to be horizontal
      # legend.position ="none", #removes legend
      plot.title = element_text(hjust=0.5),
      plot.subtitle = element_text(hjust=0.5))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=0.5, size=6), #aesthetics of the barchart
        strip.background = element_rect(fill = "#362178"),
        strip.text = element_text(color = "white", size=8, face="bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
panel.background = element_blank())+
  facet_wrap(. ~ Race, ncol=3, scales = 'free') 





```



### Traffic stops resulting in driving without license citation

Across all officer-initiated traffic stops, __non-moving traffic stops__ have the highest rates of stops that result in a citation for driving without a license. 

```{r}

df<-traffic%>%
  left_join(citation_result%>%
              mutate(person_number=as.character(person_number)))%>%
  left_join(offense_codes, by=c("offense_code"="offense_code"))%>%
  group_by(traffic_violation_type)%>%
  mutate(total=n())%>%
  filter(offense_statute=="12500(A)" |
          offense_statute=="12500(B)" |
           offense_statute=="12500(C)" |
           offense_statute=="12500(D)")%>%
  mutate(count=n(),
         rate=count/total*100)%>%
  slice(1)%>%
  select(traffic_violation_type, statute_literal_25, total, count,rate)%>%
  arrange(-rate)

# create and style flextable

names(df)<-tools::toTitleCase(gsub("_"," ",names(df))) # clean up column names for table

df<-df%>%
  rename("Statute"="Statute Literal 25")

  flextable(df) %>%
colformat_double(j = c("Rate"), digits = 1)%>%
  set_caption('Traffic Stops Resulting in Driving without License Citation by Stop Type') %>%
  theme_vanilla() %>%
  width(j=2,2,unit="in")

```

People Fresno PD perceive as Latinx and Black are stopped in traffic and given a citation for driving without a license at higher rates than other groups. The percentage of people Fresno perceived as Black who were stopped and given a citation for driving without a license is disproportionately high relative to the percentage of Black people who live in Fresno city. 

```{r}


df<-traffic%>%
  left_join(citation_result%>%
              mutate(person_number=as.character(person_number)))%>%
  left_join(offense_codes, by=c("offense_code"="offense_code"))%>%
   filter(offense_statute=="12500(A)" |
          offense_statute=="12500(B)" |
           offense_statute=="12500(C)" |
           offense_statute=="12500(D)")%>%
  mutate(total=n())%>%
  left_join(race)%>%
  group_by(nh_race)%>%
  mutate(count=n(),
         rate=count/total*100)%>%
  slice(1)%>%
  select(nh_race, offense_statute, statute_literal_25, total, count,rate)%>%
  arrange(-rate)%>%
     mutate(Race = case_when(
    nh_race == 'latinx' ~ "Latinx", 
    nh_race =='nh_white' ~ 'White',
    nh_race =='nh_black' ~ 'Black',
    nh_race == 'nh_asian' ~ 'Asian',
    nh_race == 'nh_aian' ~ 'AIAN',
    nh_race == 'nh_nhpi' ~ 'NHPI',
    nh_race == 'nh_multiracial' ~ 'Multiracial',
     nh_race == 'nh_sswana' ~ 'SWANA/SA', 
    TRUE ~ as.character(nh_race)))

# make table 

names(df)<-tools::toTitleCase(gsub("_"," ",names(df))) # clean up column names for table

df<-df%>%
  ungroup()%>%
  rename("Statute"="Statute Literal 25")%>%
  select(Race, "Statute", "Total", "Count", "Rate")

  flextable(df) %>%
colformat_double(j = c("Rate"), digits = 1)%>%
  set_caption('Traffic Stops Resulting in Driving without License Citation by Perceived Race') %>%
  theme_vanilla() %>%
  width(j=2,2,unit="in")

```

# Stop Results  {#results}

## Officer-initiated stops by stop result

```{r}

df<-stops%>%
  left_join(results)%>%
  filter(call_for_service==0)%>%
  group_by(stop_result_simple)%>%
summarise(count=n(),.groups='drop')%>%
mutate(rate=count/sum(count)*100)%>%
  arrange(-rate)

# make table 

names(df)<-tools::toTitleCase(gsub("_"," ",names(df))) # clean up column names for table

  flextable(df) %>%
colformat_double(j = c("Rate"), digits = 1)%>%
  set_caption('Officer-Initiated Stops by Stop Result') %>%
  theme_vanilla() %>%
  width(j=1, width=3, unit="in")

```


## Stops resulting in use of force {#useofforce}

Out of all people stopped during an officer-initiated stop, 99.7% of people stopped were indicated as having no instances of any use of force used towards them (5,043 total people) while 0.33% of people stopped were indicated in the data as having instances of use of force used against them (17 people). All officer-initiated incidents of use of force inflicted by Fresno PD in 2022 occurred during __traffic stops__. 

```{r}

# Create universe for use of force analysis: person and stop level data joined together and filtered for officer-initiated stops

uof<-person%>%
  mutate(person_number=as.character(person_number))%>%
  left_join(stops)%>%left_join(actions)%>%
    filter(call_for_service==0)

# Analysis

df<-uof%>%
  mutate(total=n())%>%
  group_by(use_of_force)%>%
  mutate(count=n(),
         rate=count/total*100)%>%
  slice(1)%>%
  select(use_of_force, total, count, rate)


```

```{r}

uof[,1:149] <- sapply(uof[,1:149],as.character) # convert all columns to varchar for joining

df0<-uof%>%
  filter(use_of_force==1)%>%
  left_join(person_reason, by =c("stop_id","person_number"))%>%
    mutate(Total=n())%>%
  group_by(reason, rfs_traffic_violation_code.x)%>%
  mutate(Count=n(),
         Rate=Count/Total*100)%>%
  slice(1)%>%
  select(use_of_force, reason, rfs_traffic_violation_code.x, Total, Count, Rate)%>%
  arrange(-Count)%>%
  left_join(offense_codes, by=c("rfs_traffic_violation_code.x"="offense_code"))%>%
  ungroup()%>%
  select(reason, rfs_traffic_violation_code.x, statute_literal_25, Total, Count, Rate)%>%
  rename("Traffic reason offense code"='rfs_traffic_violation_code.x',
         "Statute"="statute_literal_25")

#create and style flextable

names(df)<-tools::toTitleCase(gsub("_"," ",names(df))) # clean up column names for table

# Create table so we can look at the traffic citation reasons for UoF

  flextable(df0) %>%
colformat_double(j = c("Rate"), digits = 1)%>%
  set_caption('Use of Force by Stop Reason') %>%
  theme_vanilla() %>%
  width(j=3,2.5,unit="in")



```


Fresno PD disproportionately inflicted use of force towards people they perceived as Latinx men in 2022. Additionally, __18% of use of force incidents were towards people perceived as Asian__ despite Asian people only making up __~13% of Fresno's total population__. 

```{r}

df_gender<-uof%>%
left_join(race)%>%
  filter(use_of_force==1)%>%
group_by(nh_race, g_full)%>%
summarise(count=n(),.groups='drop')%>%
mutate(rate=count/sum(count))


# Format the gender table for visualizing 

df<-df_gender%>%
  mutate(g_full=ifelse(g_full %in% '1', 'Male',
                       ifelse(g_full %in% '2', 'Female',
                              g_full)))%>%
    mutate(nh_race=ifelse(nh_race %in% 'latinx', 'Latinx',
                       ifelse(nh_race %in% 'nh_asian', 'Asian',
                              ifelse(nh_race %in% 'nh_sswana', 'SSWANA',
                              "White"))))%>%
  rename("Perceived Gender"="g_full",
         "Perceived Race"="nh_race")%>%
  ungroup()%>%
  select(-3)%>%
  arrange("Perceived Race", -rate)

#create and style flextable

names(df)<-tools::toTitleCase(gsub("_"," ",names(df))) # clean up column names for table

  
flextable(df) %>%
colformat_double(j = c("Rate"), digits = 1)%>%
  set_caption('Use of Force by Perceived Race and Gender') %>%
  theme_vanilla() %>%
  width(j=3,2.5,unit="in")


  


```

## Stops resulting in a field interview (FI) card {#ficard}

Two officer-initiated stops by Fresno PD in 2022 resulted solely in a field interview card being completed. A total of __four people__ were stopped and given FI cards during an officer-initiated stop. All four FI cards were given to people during __traffic stops__. 

Every individual that Fresno PD stopped and completed FI cards for in 2022 was perceived by officers as __Black or Latinx__. Officers also perceived three of these individuals as male and one as female.

```{r, include=FALSE}

##Create df of all officer-initiated stops w/ stop result variables##
ois_result <- left_join(stops, person_result, by ="stop_id")%>%
  filter(call_for_service ==0) %>%
  select(stop_id, stop_result_simple, stop_result_list, everything())

##Create df to find count and rate of all stop results##
count_ois_result <- ois_result %>%
  group_by(stop_result_simple) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  mutate(Percent = Count/sum(Count) * 100) %>%
    arrange(desc(Count)) %>%               # Change column names and value                                                 displays for more legible table
  rename('Recorded Stop Result' = stop_result_simple, 
         `# of Times Recorded` = Count, 
         `% of All Stop Results` = Percent) %>%
     mutate(`% of All Stop Results` = sprintf("%.2f%%", `% of All Stop Results`))

##Table to show result counts and rates of all OIS, with FI result highlighted##
flextable(count_ois_result) %>%
  set_caption('Results of Officer-Initiated Stops by Fresno PD (2022)') %>%
  theme_vanilla() %>%
  width(width=2,unit="in") %>%
    align(j = 3, align = "right", part = "body") %>%
    align(j = 3, align = "right", part = "header") %>%
   bold(i = 8, bold = TRUE, part = "body") %>%
  color(i = 8, color = "red", part = "body")


```

```{r, include=FALSE}

##Create df that includes only FI card results and merges in stop reason data##
fi_stop_reason <- left_join(ois_result, reason, by = "stop_id") %>%
  filter(
    stop_result_simple == "field interview card completed" |
    grepl("field interview", stop_result_list, ignore.case = TRUE)) %>%
  select(stop_id, stop_result_simple, stop_reason_simple, stop_result_list, everything())    #Only 4 stops where result was FI, reason for all was TV


```

```{r, include=FALSE}

##Merge age_race_gender with FI card df##
fi_stop_race_gender <- left_join(fi_stop_reason, age_race_gender_ois, by = "stop_id")   #Df shows 5 individuals who had FI cards filed,
              # 2 Black, 2 Latinx, 1 white.
              # 3 men, 2 women

```

## Stops resulting in referral to US Department of Homeland Security ICE {#ice}

Fresno PD reported that only one of their stops in 2022 resulted in a person being referred to ICE. The one individual referred to ICE by Fresno PD in 2022 was stopped for a traffic violation.

The one individual referred to ICE by Fresno PD in 2022 was perceived by officers as __a Latinx male__.


```{r, include=FALSE}

##count_ois_result shows us 0 under simple results, 
##query to see how many times it is listed under multiple results##
sum(grepl("homeland security", ois_result$stop_result_list, ignore.case = TRUE)) #only 1 total referral to ICE

```

```{r, include=FALSE}
##Create df for one ICE result and merge reason##
ice_reason <- left_join(ois_result, reason, by = "stop_id") %>%
  filter(grepl("homeland security", stop_result_list, ignore.case = TRUE)) %>%
  select(stop_id, stop_result_simple, stop_reason_simple, stop_result_list, everything())

```

```{r, include=FALSE}

##Merge in race and gender variables for one observation##
ice_race_gender <- left_join(ice_reason, age_race_gender_ois, by = "stop_id")

```

## Stops that resulted in a search that yielded no contraband {#contraband}

In 2022, Fresno PD made 68 searches during officer-initiated traffic stops and searched 59 people. Of the 59 people Fresno PD searched in traffic, 32 people were perceived as Latinx and 9 people were perceived as Black.  

Out of the 68 searches made in traffic, 50 of those searches did not yield any contraband. In other words, 73.5% searches made in traffic did not prevent any crime.

```{r}

df<-person_action%>%
  left_join(person_reason%>%
              mutate(person_number=as.numeric(person_number)))%>%
  left_join(stops)%>%
  filter(call_for_service==0)%>%
  filter(ads_search_person==1 & reason == "Traffic violation")%>%
  left_join(race%>%mutate(person_number=as.numeric(person_number)))

>>>>>>> ef0697534eb84c3dbe3b926af4394eb5d26fb9c1
# explore stops resulting in a search for some general counts

df<-search%>%
  left_join(stops)%>%
  left_join(reason)%>%
  filter(stop_reason_simple=="Traffic violation")%>%
  filter(call_for_service==0)%>%
  group_by(search)%>%
  summarise(Count = n(),
            .groups = 'drop')%>%
  mutate(percent=Count/sum(Count)*100)

# explore stops resulting in a search and contraband

df_contra<-search%>%
  left_join(stops)%>%
  left_join(reason)%>%
  filter(stop_reason_simple=="Traffic violation")%>%
  filter(call_for_service==0)%>%
  group_by(search,contraband_found)%>%
  summarise(Count = n(),
            .groups = 'drop')



```

A total of 80 people were searched by Fresno PD in 2022. Of those 80 people searched by Fresno PD, __16 people searched were perceived as Black by officers, or 20% of all searches.__ More than half of all searches made by Fresno PD were of people perceived as Latinx.

```{r}

# read in raw data from postgres:

person_search<-dbGetQuery(con, "SELECT * FROM  cadoj_ripa_fresno_2022
WHERE ads_search_person	= 1 AND call_for_service=0")%>%
  rename("stop_id"="doj_record_id")

# analyze just searches by race -------------

df<-person_search%>%
  left_join(person_reason%>%
              mutate(person_number=as.numeric(person_number)),
             by=c("stop_id", "person_number"))%>%
  left_join(race%>%mutate(person_number=as.numeric(person_number)),
            by=c("stop_id", "person_number"))%>%
  filter(reason == "Traffic violation")%>%
  mutate(Total=n())%>%
  group_by(nh_race)%>%
  mutate(Count=n(),
         Rate=Count/Total*100)%>%
  slice(1)%>%
  ungroup()%>%
  mutate(Race = case_when(
    nh_race == 'latinx' ~ "Latinx", 
    nh_race =='nh_white' ~ 'White',
    nh_race =='nh_black' ~ 'Black',
    nh_race == 'nh_asian' ~ 'Asian',
    nh_race == 'aian' ~ 'AIAN',
    nh_race == 'nhpi' ~ 'NHPI',
    nh_race == 'nh_multiracial' ~ 'Multiracial',
     nh_race == 'nh_sswana' ~ 'SWANA/SA', 
    TRUE ~ as.character(nh_race)))%>%
  select(Race, Total, Count, Rate)%>%
  arrange(-Rate)

# create table of just searches by race------------

#create and style flextable
names(df)<-tools::toTitleCase(gsub("_"," ",names(df))) # clean up column names for table


  flextable(df) %>%
 colformat_double(j = c("Rate"), digits = 1)%>%
  set_caption('Fresno PD Searches by Perceived Race') %>%
  theme_vanilla() 

```

Despite Fresno PD disproportionately stopping and searching people perceived as Black and Latinx in traffic stops, these groups also have the highest rates of searches that yield no contraband (also known as hit rates). In contrast, out of all the searches Fresno PD conducted towards people perceived as White during traffic stops, only 2 of those searches yielded no contraband. In other words, __the majority of searches of people that yield contraband during traffic stops is of people perceived as White.__ 

```{r}

# analyze HIT RATES by race:  searches that yielded no contraband--------------------

df_contra_person<-person_search%>%
  left_join(person_reason%>%
              mutate(person_number=as.numeric(person_number)),
             by=c("stop_id", "person_number"))%>%
  left_join(race%>%mutate(person_number=as.numeric(person_number)),
            by=c("stop_id", "person_number"))%>%
  filter(reason == "Traffic violation")%>%
  group_by(nh_race)%>%
  mutate(Total=n())%>%
  filter(ced_none_contraband==0)%>%
  mutate(Count=n(),
         Rate=Count/Total*100)%>%
  slice(1)%>%
    ungroup()%>%
    mutate(Race = case_when(
    nh_race == 'latinx' ~ "Latinx", 
    nh_race =='nh_white' ~ 'White',
    nh_race =='nh_black' ~ 'Black',
    nh_race == 'nh_asian' ~ 'Asian',
    nh_race == 'aian' ~ 'AIAN',
    nh_race == 'nhpi' ~ 'NHPI',
    nh_race == 'nh_multiracial' ~ 'Multiracial',
     nh_race == 'nh_sswana' ~ 'SWANA/SA', 
    TRUE ~ as.character(nh_race)))%>%
  select(Race, Total, Count, Rate)%>%
  arrange(-Rate)
  
  #create and style flextable
names(df_contra_person)<-tools::toTitleCase(gsub("_"," ",names(df_contra_person))) # clean up column names for table

  flextable(df_contra_person) %>%
 colformat_double(j = c("Rate"), digits = 1)%>%
  set_caption('Fresno PD Searches Yielding No Contraband by Perceived Race') %>%
  theme_vanilla() 


```

# Time spent {#time}

The following analyzes the total and percent of time Fresno PD spent on certain stops to explore potential waste in public resources. We focus on officer-initiated stops. Overall, Fresno PD spent about __74.1% of their stop time on officer-initiated stops__ though stops for calls for service on average took longer (56 minutes on average per call for service stop compared to 9 minutes on average for officer-initiated).

## Time spent on stops by stop reason {#time-reason}

In 2022, Fresno PD spent nearly __98% of their stop time on stops originally made for traffic violations.__ While each of these stops took less time on average compared to other stop types, they made up the largest share of Fresno PD's time.

```{r, include=FALSE}

# join data frames
time_reason_df<-stops%>%left_join(reason)

# get quick tallies of OIS vs. cfs
time_ois<-time_reason_df%>%group_by(call_for_service)%>%
  summarise(count_time=sum(stop_duration),
            avg_time=mean(stop_duration),
            median_time=median(stop_duration),
            .groups='drop')%>%
           mutate(percent_time=count_time/sum(count_time))

# time spent on stop reason for ois
time_reason_ois<-time_reason_df%>%filter(call_for_service==0)%>%
  group_by(stop_reason_simple)%>%
  summarise(total_time=sum(stop_duration),
            average_time=round(mean(stop_duration),1),
            median_time=median(stop_duration),
            .groups='drop')%>%
           mutate(percent_time=round(total_time/sum(total_time)*100,1))%>%
  arrange(desc(percent_time))%>%
  rename(stop_reason=stop_reason_simple)


#create and style flextable
names(time_reason_ois)<-tools::toTitleCase(gsub("_"," ",names(time_reason_ois))) # clean up column names for table

  flextable(time_reason_ois) %>%
# colformat_double(j = c("Rate"), digits = 1)%>%
  set_caption('Time Spent in Minutes on Officer-initiated Stops by Stop Reason') %>%
  theme_vanilla() 



```

## Time spent on traffic stops by stop result {#time-result}

The large majority of time spent on traffic stops is spent on stops only resulting in a citation for infraction. Nearly 73% of minutes Fresno PD spent on traffic stops were on stops resulting in a citation for infraction. On average, each of these stops only takes 8 minutes.

```{r, include=FALSE}

# join data frames
time_result_df<-time_reason_df%>%left_join(results)

# time spent by result for ois and traffic
time_result_ois<-time_result_df%>%filter(call_for_service==0 & stop_reason_simple=='Traffic violation')%>%
  group_by(stop_result_simple)%>%
  summarise(total_time=sum(stop_duration),
            average_time=round(mean(stop_duration),1),
            median_time=median(stop_duration),
            .groups='drop')%>%
           mutate(percent_time=round(total_time/sum(total_time)*100,1))%>%
  arrange(desc(percent_time))%>%
  rename(stop_result=stop_result_simple)%>%
  mutate(stop_result=tools::toTitleCase(stop_result))

#create and style flextable
names(time_result_ois)<-tools::toTitleCase(gsub("_"," ",names(time_result_ois))) # clean up column names for table

  flextable(time_result_ois) %>%
# colformat_double(j = c("Rate"), digits = 1)%>%
  set_caption('Time Spent in Minutes on Traffic Stops by Stop Result') %>%
  theme_vanilla()%>%   
    width(width=2,unit="in")



```

## Time spent on traffic stops by race {#time-race}

On average, Fresno PD spent slightly longer on traffic stops of Black and Latinx people (about 1 minute longer looking at the median time or quartile 3 time). Quartile 1 represents the 25th percentile of stop duration, meaning that 25% stops in that racial group took less than that specified time. Quartile 3 represents the 75th percentile of stop duration, meaning that 75% stops in that racial group took less than that specified time.

Stops of White individuals in Fresno only took 1 in 4 minutes of Fresno PD's stop time on traffic violations.

```{r, include=FALSE}

# join data frames
time_race_df<-time_result_df%>%left_join(race)

# time spent by race for traffic
time_race_traffic<-time_race_df%>%filter(call_for_service==0 & stop_reason_simple=='Traffic violation')%>%
  group_by(nh_race)%>%
  summarise(total_time=sum(stop_duration),
            average_time=round(mean(stop_duration),1),
            median_time=median(stop_duration),
            count=n(),
            quartile_1_time=quantile(stop_duration,probs=c(.25)),
           quartile_3_time=quantile(stop_duration,probs=c(.75)),
            .groups='drop')%>%
           mutate(percent_time=round(total_time/sum(total_time)*100,1))%>%
  arrange(desc(percent_time))

# clean up 
time_race_traffic_df<-time_race_traffic%>%
 mutate(Race = case_when(
    nh_race == 'latinx' ~ "Latinx", 
    nh_race =='nh_white' ~ 'White',
    nh_race =='nh_black' ~ 'Black',
    nh_race == 'nh_asian' ~ 'Asian',
    nh_race == 'aian' ~ 'AIAN',
    nh_race == 'nhpi' ~ 'NHPI',
    nh_race == 'nh_multiracial' ~ 'Multiracial',
     nh_race == 'nh_sswana' ~ 'SWANA/SA', 
    TRUE ~ as.character(nh_race)))%>%
  filter(count>5)%>%
  select(-count,-nh_race)%>%
  select(Race,total_time,percent_time,everything())
    
#create and style flextable
names(time_race_traffic_df)<-tools::toTitleCase(gsub("_"," ",names(time_race_traffic_df))) # clean up column names for table

  flextable(time_race_traffic_df) %>%
# colformat_double(j = c("Rate"), digits = 1)%>%
  set_caption('Time Spent in Minutes on Traffic Stops by Race') %>%
  theme_vanilla() 
    # width(width=2,unit="in")



```


